---
title: "PD & EL plot walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PD & EL plot walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(magrittr)
```

# Why this vignette

The `pd-el-integration` vignette walks the methodology end to end: every step,
every formula, the full bank-risk story. This one is the **visual reference**.
Each section covers one plot or table: what it shows, what to look at first,
common misreads, and when to reach for it instead of the others.

For the maths — Z-score method, EAD-weighting, Basel IRB alignment, EL sign
convention — see [`pd-el-integration.Rmd`](pd-el-integration.html).

The seven artefacts:

| Tag | Function | Family |
|----|----------|--------|
| P1 | `pipeline_crispy_pd_integration_bars()` | PD level |
| P2 | `pipeline_crispy_el_adjustment_bars()` | EL delta |
| P3a | `pipeline_crispy_pd_kpi_table()` | KPI summary |
| P3b | `pipeline_crispy_el_kpi_table()` | KPI summary |
| P4 | `pipeline_crispy_el_sector_breakdown_table()` | Sector breakdown |
| N1 | `pipeline_crispy_pd_method_comparison()` | Method choice |
| N2 | `pipeline_crispy_pd_waterfall()` | PD decomposition |

# Setup (bundled testdata)

```{r}
assets_testdata    <- read.csv(system.file("testdata", "assets_testdata.csv",    package = "trisk.model"))
scenarios_testdata <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model"))
fin_testdata       <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model"))
carbon_testdata    <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model"))
portfolio_ids_internal_pd <- read.csv(system.file("testdata", "portfolio_ids_internal_pd_testdata.csv",
                                                  package = "trisk.analysis"))

analysis_data <- run_trisk_on_portfolio(
  assets_data       = assets_testdata,
  scenarios_data    = scenarios_testdata,
  financial_data    = fin_testdata,
  carbon_data       = carbon_testdata,
  portfolio_data    = portfolio_ids_internal_pd,
  baseline_scenario = "NGFS2023GCAM_CP",
  target_scenario   = "NGFS2023GCAM_NZ2050"
)

internal_pd_lookup <- portfolio_ids_internal_pd[, c("company_id", "internal_pd")]

# Z-score is the default integration method; used for all PD plots below.
result_zs <- integrate_pd(analysis_data, internal_pd = internal_pd_lookup)

# EL pipeline needs derived columns + an internal EL lookup.
analysis_data_el <- compute_analysis_metrics(analysis_data)
internal_el_lookup <- merge(
  analysis_data_el[, c("company_id", "exposure_value_usd", "loss_given_default")],
  internal_pd_lookup,
  by = "company_id"
)
internal_el_lookup$internal_el <-
  internal_el_lookup$exposure_value_usd *
  internal_el_lookup$loss_given_default *
  internal_el_lookup$internal_pd

result_el <- integrate_el(analysis_data_el,
                          internal_el = internal_el_lookup[, c("company_id", "internal_el")])
```

# P1 — PD integration bars

Four bars per facet group: **Internal PD** (the bank's own number), **TRISK
Baseline PD** (no shock), **TRISK Shock PD** (with the transition shock), and
**TRISK-Adjusted PD** (Internal moved by the TRISK delta). The first thing to
look at is the gap between **Internal** and **TRISK-Adjusted** — that's how much
your number changes once climate is layered in. The TRISK Baseline and Shock
bars sit alongside for context: they tell you *where the model itself moved*,
which is what the integration method translates into a shift on your internal
number.

```{r p1-pd-integration-bars, fig.width=7, fig.height=4}
pipeline_crispy_pd_integration_bars(result_zs)
```

> **When to use.** First plot to show a bank counterparty. Answers "what
> happens to my PDs if I integrate TRISK?" at sector resolution. Switch to
> `granularity = "firm"` when sector aggregation hides material within-sector
> heterogeneity (sparse portfolios where a few firms drive everything). Use
> `scale = "pseudo_log"` if Merton underflow leaves baseline PDs near zero and
> linear scaling collapses everything against the axis.

# P2 — EL adjustment bars

Horizontal bars per sector showing the **signed EL delta**: TRISK-Adjusted EL
minus Internal EL. Red bars (positive) mean integration says you should expect
**more loss** than your own model predicted; green bars (negative) mean less
loss. The EL levels themselves aren't on this plot — only the adjustment — so
this is the right view when the conversation is "how much does integration
change my numbers" rather than "what are the absolute levels."

```{r p2-el-adjustment-bars, fig.width=7, fig.height=4}
pipeline_crispy_el_adjustment_bars(result_el)
```

> **When to use.** When the question is sign and magnitude of integration
> impact on EL by sector — "where does TRISK think you're under-reserving?"
> Pair with **P4** (sector breakdown) to show levels and direction together.

# P3a — PD KPI table

One-row summary of the PD integration: portfolio Internal PD, portfolio
TRISK-Adjusted PD, the delta, and direction. The number on the right is what
goes into a one-page memo or an executive deck.

```{r p3a-pd-kpi-table}
pipeline_crispy_pd_kpi_table(result_zs$aggregate)
```

> **When to use.** Top of a report or first slide of a deck. Strips out every
> sector and firm — useful when the question is portfolio-level only. Don't
> use as the *only* PD view in a methodology document: it hides every dimension
> the integration cares about (sector, firm, method).

# P3b — EL KPI table

One-row summary of the EL integration, including the **bps metric**
(`EL / EAD`, expressed in basis points). The bps number is the line that
travels best across audiences — it's scale-free, directly comparable to risk
appetite limits, and the same unit regulators use for IRB capital math.

```{r p3b-el-kpi-table}
pipeline_crispy_el_kpi_table(result_el$aggregate)
```

> **When to use.** Same as P3a but for EL. If you only have room for one
> headline number per side, the **bps delta** is the best candidate — it
> compresses absolute level, exposure, and integration impact into one figure.

# P4 — EL sector breakdown table

The portfolio breakdown that sits between P2 (deltas only) and P3b (one
number). Each row is a sector with: exposure, internal EL, TRISK-adjusted EL,
delta, direction arrow, and EL/EAD in bps. Reading top to bottom tells you
which sectors carry the integration story.

```{r p4-el-sector-breakdown}
pipeline_crispy_el_sector_breakdown_table(result_el$portfolio)
```

> **When to use.** The reference table for any EL-focused discussion. Shows
> levels and direction together — P2 only shows deltas, P3b collapses sectors.
> The bps column makes sectors with very different exposure magnitudes
> directly comparable.

# N1 — PD method comparison

The three integration methods (`"absolute"`, `"relative"`, `"zscore"`) overlaid
on the same input. The Internal PD shows as a tick; each method's
TRISK-adjusted PD shows as a coloured shape on the same row. Wide spread
between method shapes means the bank's choice of integration method moves the
number a lot — narrow spread means it doesn't matter which method you pick,
the integration tells the same story.

```{r n1-pd-method-comparison, fig.width=7, fig.height=4}
pipeline_crispy_pd_method_comparison(analysis_data,
                                     internal_pd = internal_pd_lookup)
```

> **When to use.** When the methodology choice is on the table — internal
> committee review, regulator pushback on method selection, or before
> committing to a single method for a client engagement. If the three shapes
> sit on top of each other, picking a method is a non-decision; if they
> spread, that's a real call that needs justification.
>
> Like P1, this plot exposes `granularity = "firm"` and `scale = "pseudo_log"`
> for sparse portfolios with near-zero baseline PDs.

# N2 — PD waterfall

Per-sector decomposition: **Internal PD → signed Adjustment → Adjusted PD**.
The middle bar's sign and fill make the integration impact unmissable: red
means the adjustment worsens risk, green means it improves it. Reading left
to right gives the magnitude story at a glance — useful when the audience
needs the "before, change, after" structure rather than four separate bars.

```{r n2-pd-waterfall, fig.width=7, fig.height=4}
pipeline_crispy_pd_waterfall(result_zs)
```

> **When to use.** When P1's four-bar grouped layout is too dense for the
> audience — non-quantitative stakeholders read waterfalls faster than
> grouped bars. Also the default choice when the message is "this is how
> much integration moves PD" rather than "here are the four reference
> levels." Pair P1 + N2 when both questions are in scope.

# How they fit together

A typical reading order for a bank-facing report:

1. **P3a + P3b** — headline numbers, one slide.
2. **N2** (PD waterfall) and **P2** (EL adjustment) — the *change* story.
3. **P1** and **P4** — full reference levels, sector by sector.
4. **N1** — only if the methodology choice is being questioned.

For deeper math, every chart cross-references back to
[`pd-el-integration.Rmd`](pd-el-integration.html) section by section.
