---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# What is trisk.analysis

`trisk.analysis` measures **climate transition risk** at the level of individual
companies and the portfolios built from them — wherever transition risk has to be
placed on a book of exposures, whether by banks, central banks and supervisors, or
asset managers.

The underlying TRISK engine (`trisk.model`) is an asset-level transition-risk
stress test: it feeds climate scenarios into a microeconomic model of the firm and
uses company- and asset-level data to value how scenario-driven declines in
high-carbon production and stranded-asset exposure adjust a company's value
relative to a business-as-usual baseline. TRISK turns those scenario assumptions
into shocked **net present values (NPVs)** and **probabilities of default (PDs)**.

This package wraps that engine so you can run it directly on a portfolio, allocate
the shocks onto your own exposures — loans or equity holdings — and blend them with
your internal PD/EL estimates. The worked vignettes below follow a **bank**
credit-risk workflow; other institution types can adapt the same steps.

# Reading path

The worked vignettes follow a bank credit-risk workflow. Work through them in
this order:

1. `vignette("bank_1_portfolio-analysis")` — the bank entry point: the five
   inputs, setting up your `trisk_inputs/` data folder, and a first end-to-end run.
2. `vignette("bank_2_inputs-and-outputs")` — what TRISK consumes (assets,
   scenarios, financial features, carbon prices) and what it produces (NPV and PD
   results).
3. `vignette("bank_3_run-on-a-portfolio")` — run TRISK on a loan book and map the
   shocks onto your exposures.
4. `vignette("bank_4_sensitivity-analysis")` — sweep scenario assumptions to see
   how results move across baselines and shock scenarios.
5. `vignette("bank_5_pd-el-integration")` — blend TRISK shocks with your own
   PD/EL to get climate-adjusted expected loss.

# A first run

The smallest end-to-end run: load the four bundled `trisk.model` inputs and a
bundled portfolio, then call `run_trisk_on_simple_portfolio()`.

```{r setup}
library(trisk.analysis)
```

```{r first-run}
assets <- read.csv(system.file("testdata", "assets_testdata.csv", package = "trisk.model", mustWork = TRUE))
scenarios <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model", mustWork = TRUE))
financial_features <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model", mustWork = TRUE))
ngfs_carbon_price <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model", mustWork = TRUE))

portfolio <- read.csv(
  system.file("testdata", "portfolio_ids_internal_pd_testdata.csv", package = "trisk.analysis", mustWork = TRUE)
)

results <- run_trisk_on_simple_portfolio(
  assets_data = assets,
  scenarios_data = scenarios,
  financial_data = financial_features,
  carbon_data = ngfs_carbon_price,
  portfolio_data = portfolio,
  baseline_scenario = "NGFS2023GCAM_CP",
  target_scenario = "NGFS2023GCAM_NZ2050",
  scenario_geography = "Global"
)

head(results$portfolio_results)
```

Each input row comes back with its climate-shocked exposure loss and expected
loss alongside the original exposure.

## All model parameters

`run_trisk_on_simple_portfolio()` forwards any extra arguments to the TRISK
engine (`trisk.model::run_trisk_model()`). Only the scenarios above are required;
every other argument has a default, shown here with its meaning. Set them to match
your own cost-of-capital assumptions and the prevailing market curve at your
analysis date.

```{r all-parameters, eval = FALSE}
results <- run_trisk_on_simple_portfolio(
  assets_data        = assets,
  scenarios_data     = scenarios,
  financial_data     = financial_features,
  carbon_data        = ngfs_carbon_price,
  portfolio_data     = portfolio,
  baseline_scenario  = "NGFS2023GCAM_CP",       # business-as-usual scenario
  target_scenario    = "NGFS2023GCAM_NZ2050",   # shock (policy) scenario
  scenario_geography = "Global",                # region(s) to compute results for
  carbon_price_model = "no_carbon_tax",         # NGFS carbon-price pathway ("no_carbon_tax" to skip)
  risk_free_rate     = 0.045,                   # risk-free rate (Merton PD model)
  discount_rate      = 0.09,                    # DCF discount rate on dividends
  growth_rate        = 0.03,                    # terminal growth rate of profits
  div_netprofit_prop_coef = 1,                  # dividend pass-through coefficient
  shock_year         = 2030,                    # year the policy shock is applied
  market_passthrough = 0                        # firm's ability to pass carbon cost to consumers
)
```

See `vignette("bank_2_inputs-and-outputs")` for fuller definitions of each
parameter.

# Where to go next

Start with `vignette("bank_1_portfolio-analysis")` to set up your inputs and run
TRISK on your own portfolio.
