---
title: "1. Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# What is trisk.analysis

`trisk.analysis` helps a bank credit-risk analyst measure **climate transition
risk** on a loan book. The underlying TRISK engine (`trisk.model`) re-prices the
companies you lend to under different climate scenarios: as carbon prices rise
and demand shifts away from high-carbon technologies, some borrowers lose value
and become more likely to default. TRISK turns those scenario assumptions into
shocked **net present values (NPVs)** and **probabilities of default (PDs)**.
This package wraps that engine so you can run it directly on a portfolio,
allocate the shocks to your exposures, and blend them with your own internal
PD/EL estimates.

# Reading path

Work through the vignettes in this order:

1. `vignette("inputs-and-outputs")` — what TRISK consumes (assets, scenarios,
   financial features, carbon prices) and what it produces (NPV and PD results).
2. `vignette("run-on-a-portfolio")` — run TRISK on a bank loan book and map the
   shocks onto your exposures.
3. `vignette("pd-el-integration")` — blend TRISK shocks with the bank's own
   PD/EL to get climate-adjusted expected loss.
4. `vignette("sensitivity-analysis")` — sweep scenario assumptions to see how
   results move across baselines and shock scenarios.

# Hello world

The smallest end-to-end run: load the four bundled `trisk.model` inputs and a
bundled portfolio, then call `run_trisk_on_simple_portfolio()`.

```{r setup}
library(trisk.analysis)
```

```{r hello-world}
assets <- read.csv(system.file("testdata", "assets_testdata.csv", package = "trisk.model", mustWork = TRUE))
scenarios <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model", mustWork = TRUE))
financial_features <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model", mustWork = TRUE))
ngfs_carbon_price <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model", mustWork = TRUE))

portfolio <- read.csv(
  system.file("testdata", "portfolio_ids_internal_pd_testdata.csv", package = "trisk.analysis", mustWork = TRUE)
)

results <- run_trisk_on_simple_portfolio(
  assets_data = assets,
  scenarios_data = scenarios,
  financial_data = financial_features,
  carbon_data = ngfs_carbon_price,
  portfolio_data = portfolio,
  baseline_scenario = "NGFS2023GCAM_CP",
  target_scenario = "NGFS2023GCAM_NZ2050",
  scenario_geography = "Global"
)

head(results$portfolio_results)
```

Each input row comes back with its climate-shocked exposure loss and expected
loss alongside the original exposure.

# Where to go next

Start with `vignette("inputs-and-outputs")` to understand the data TRISK needs
before running it on your own portfolio.
