---
title: "Bank workflow 4: PD and EL integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bank workflow 4: PD and EL integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(magrittr)
```

# PD and EL Integration

## Overview

TRISK recomputes probability of default (PD) from a Merton structural credit
model under a climate-transition shock. That model PD is **not** directly
comparable to the internal PD your institution already carries: levels differ
by calibration, rating philosophy, and point-in-time vs through-the-cycle
treatment. Only the *baseline-to-shock change* carries portable meaning.

`integrate_pd()` and `integrate_el()` translate that TRISK shift onto your
own internal scale, using one of three methods (`zscore`, the default;
`absolute`; `relative`). This vignette covers the maths, a worked numeric
example, the input schema, the seven pipeline plots/tables, and the
limitations a credit-risk analyst should keep in mind.

Audience: an analyst who already knows PD / EL / IFRS-9 but is new to TRISK.

## Inputs

> **Input data — where your data goes.** TRISK needs **five inputs**: four that
> describe the world — **assets**, **scenarios**, **NGFS carbon price** and
> **financial features** — plus your **portfolio**. The main portfolio file is
> **`portfolio_ids`** (matched by `company_id`); `portfolio_names` and
> `portfolio_countries` are options. This workflow uses the
> `portfolio_ids_internal_pd` variant (the same file plus an `internal_pd`
> column). **The CSVs loaded below are placeholders** (bundled samples) — replace
> them with your own files. See
> [Inputs and outputs](bank_1_inputs-and-outputs.html) for `setup_trisk_inputs()`
> and the `trisk_inputs/` folder convention.

### Portfolio file schema

Integration needs the bank's own PD per exposure, so it runs on the augmented
portfolio file `portfolio_ids_internal_pd_testdata.csv` — the classic
`portfolio_ids_testdata.csv` plus a mandatory `internal_pd` column.

| Column              | Type    | Meaning                                              |
|---------------------|---------|------------------------------------------------------|
| `company_id`        | int     | Counterparty key; joins TRISK output to internal PD. |
| `company_name`      | chr     | Display label (may be `NA`).                         |
| `sector`            | chr     | TRISK sector (e.g. `Oil&Gas`, `Coal`, `Power`).      |
| `technology`        | chr     | Production technology within the sector.             |
| `country_iso2`      | chr     | ISO-2 country code of the asset.                     |
| `exposure_value_usd`| dbl     | Exposure / outstanding, USD. Used as the EAD basis.  |
| `term`              | int     | Loan term in years.                                  |
| `loss_given_default`| dbl     | LGD as a fraction in `[0, 1]`.                        |
| `internal_pd`       | dbl     | Bank's own PD per exposure, in `[0, 1]`. **Mandatory.** |

`internal_pd` is mandatory because integration is meaningless without a bank
number to integrate the TRISK shift into — the setup below fails loudly if it
is missing or out of range. Internal EL is **derived**, not stored:
`internal_el = EAD * LGD * internal_pd` (see the EL section).

```{r}
assets_testdata    <- read.csv(system.file("testdata", "assets_testdata.csv",    package = "trisk.model"))
scenarios_testdata <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model"))
fin_testdata       <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model"))
carbon_testdata    <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model"))
portfolio_ids_internal_pd <- read.csv(system.file("testdata", "portfolio_ids_internal_pd_testdata.csv",
                                                  package = "trisk.analysis"))

stopifnot(
  "Portfolio file must include an `internal_pd` column per exposure." =
    "internal_pd" %in% colnames(portfolio_ids_internal_pd),
  "`internal_pd` values must be numeric in [0, 1]." =
    is.numeric(portfolio_ids_internal_pd$internal_pd) &&
      all(portfolio_ids_internal_pd$internal_pd >= 0 &
          portfolio_ids_internal_pd$internal_pd <= 1, na.rm = TRUE)
)
```

### The three integration methods

Let `p_int` be the internal PD, `p_base` the TRISK baseline PD, and
`p_shock` the TRISK shock PD for a given exposure. Each method returns an
*adjusted PD* `p_adj`, then `integrate_pd()` clips it to `[0, 1]`.

**Absolute** — add the raw PD shift onto the internal PD:

```
p_adj = p_int + (p_shock - p_base)
```

- *When useful:* the simplest, most transparent option, and safe when baseline
  PDs can be zero (no division). Trade-off: large shifts can push `p_adj` outside
  a sensible range, where the final `[0, 1]` clip then bites.

**Relative** — scale the internal PD by the proportional shift:

```
p_adj = p_int * (1 + (p_shock - p_base) / p_base)
```

- *When useful:* when the *proportional* move matters more than the absolute one
  — e.g. across exposures whose internal PD levels differ widely. Avoid it on
  zero-baseline rows: when `p_base = 0` it returns `p_int` unchanged and the shock
  signal is lost, so prefer `absolute` or `zscore` there.

**Z-score (default)** — recombine the three PDs in normal-quantile (probit)
space, using R's `qnorm()` (the standard-normal quantile / inverse CDF) and
`pnorm()` (the CDF), with each PD clipped to `[zscore_floor, zscore_cap]` before
`qnorm()`:

```
p_adj = pnorm( qnorm(p_int) + qnorm(p_shock) - qnorm(p_base) )
```

- *When useful:* the recommended default. It is zero-safe (via clipping),
  preserves the non-linear compression of PDs near the distribution tails, and
  behaves sensibly on sparse portfolios where Merton inputs drive some baseline
  PDs to underflow.

This is a **probit / Merton-style recombination of default thresholds**, not
the Basel IRB risk-weight formula. It adds the TRISK shock-minus-baseline
*distance-to-default* shift to the internal PD's implied default threshold,
then maps back to a probability. It does **not** invoke the Basel
asset-correlation / Vasicek conditional-PD capital formula — there is no
correlation parameter and no supervisory confidence level here.

The `zscore_floor` (default `1e-4`) and `zscore_cap` (default `1 - 1e-4`)
parameters bound the PDs before `qnorm()`, since `qnorm(0) = -Inf` and
`qnorm(1) = +Inf` would otherwise propagate infinities. Widen the floor toward
zero only if your internal PDs are reliably calibrated that low; tighten it to
de-sensitise the tails.

> **Watch the floor on underflow-prone books.** When a baseline (and shock) PD
> sits **below** `zscore_floor`, both clip to the floor and `qnorm(shock) -
> qnorm(baseline) = 0` — the shock signal is erased and the overlay reflects the
> clip bound, not the model. Merton baselines for investment-grade or
> short-horizon names routinely underflow well below `1e-4`. `integrate_pd()` /
> `integrate_el()` now report `aggregate$zscore_clipped_share` (the fraction of
> rows clipped) and emit a warning when most of the book is clipped. If that
> share is high, prefer `method = "absolute"`, which adds the raw PD shift and is
> immune to underflow.

### Worked example: one row through z-score

Take a single Oil&Gas exposure with `internal_pd = 0.025`, and suppose TRISK
returns `pd_baseline = 0.020` and `pd_shock = 0.060` for it. None of the three
values hit the clip bounds, so:

```{r}
z_int  <- qnorm(0.025)   # internal
z_base <- qnorm(0.020)   # TRISK baseline
z_shk  <- qnorm(0.060)   # TRISK shock
p_adj  <- pnorm(z_int + z_shk - z_base)
round(c(z_internal = z_int, z_baseline = z_base, z_shock = z_shk,
        adjusted_pd = p_adj, adjustment_pp = (p_adj - 0.025) * 100), 4)
```

The probit shift `z_shk - z_base` is positive (the shock raises model PD), so
the recombined internal PD rises above 0.025 — a worsening, which the plots
below render red.

## Minimal example

### Run TRISK on the portfolio

The `internal_pd` column rides along on the standard portfolio schema; we keep
it as a separate lookup table to feed into `integrate_pd()`. We run the bank
entry point — `run_trisk_on_simple_portfolio()` — on the company-level loan book
and use its per-technology detail as the analysis frame:

```{r}
portfolio_simple <- portfolio_ids_internal_pd[, c(
  "company_id", "company_name", "exposure_value_usd", "term", "loss_given_default"
)]

simple_results <- run_trisk_on_simple_portfolio(
  assets_data        = assets_testdata,
  scenarios_data     = scenarios_testdata,
  financial_data     = fin_testdata,
  carbon_data        = carbon_testdata,
  portfolio_data     = portfolio_simple,
  baseline_scenario  = "NGFS2023GCAM_CP",
  target_scenario    = "NGFS2023GCAM_NZ2050",
  scenario_geography = "Global"
)
# Expose the NPV-share-allocated exposure as `exposure_value_usd` so the
# EAD-weighted plots and aggregates can use it (the simple runner stores the
# allocated exposure as `exposure_at_default`).
analysis_data <- simple_results$portfolio_results_tech_detail |>
  dplyr::mutate(exposure_value_usd = exposure_at_default)

# The internal_pd lookup used throughout the rest of the vignette. TRISK returns
# company_id as character, so coerce the lookup key to match and keep the join
# on like types (read.csv reads it as integer).
internal_pd_lookup <- portfolio_ids_internal_pd[, c("company_id", "internal_pd")]
internal_pd_lookup$company_id <- as.character(internal_pd_lookup$company_id)
```

```{r echo=FALSE}
knitr::kable(head(analysis_data[, c("company_id", "sector", "technology", "term",
                                    "pd_baseline", "pd_shock",
                                    "net_present_value_baseline", "net_present_value_shock")])) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "300px")
```

#### Reproducibility: the run audit trail

The runner attaches a `trisk_run_meta` attribute to its output — the scenario
pair, `run_id`, every argument forwarded to the TRISK model, and the versions of
`trisk.analysis` and `trisk.model`. Capture it alongside your results so a run
can be reproduced and defended in model validation:

```{r}
run_meta <- attr(simple_results, "trisk_run_meta")
str(run_meta)
```

### Integrate PD — three methods

```{r}
result_abs <- integrate_pd(analysis_data,
                           internal_pd = internal_pd_lookup,
                           method      = "absolute")
result_rel <- integrate_pd(analysis_data,
                           internal_pd = internal_pd_lookup,
                           method      = "relative")
# zscore is the default; explicit here for clarity.
result_zs  <- integrate_pd(analysis_data,
                           internal_pd = internal_pd_lookup,
                           method      = "zscore")
```

```{r echo=FALSE}
knitr::kable(head(result_zs$portfolio[, c("company_id", "sector", "internal_pd",
                                          "pd_baseline", "pd_shock",
                                          "trisk_adjusted_pd", "pd_adjustment")])) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "300px")
```

You can also override internal PDs on the fly — pass a numeric vector of length
`nrow(analysis_data)` (or an alternate lookup) for a sanity-check stress:

```{r}
flat_internal <- rep(0.03, nrow(analysis_data))
result_custom <- integrate_pd(analysis_data,
                              internal_pd = flat_internal,
                              method      = "zscore")
```

### Integrate EL

`integrate_el()` needs `expected_loss_baseline` / `expected_loss_shock`, which the
simple runner already emits in its per-technology detail (`EAD * LGD * PD`). For
the bank's internal EL we use the same identity on the company-level loan book:
`internal_el = EAD * LGD * internal_pd`.

```{r}
internal_el_lookup <- portfolio_ids_internal_pd[, c(
  "company_id", "exposure_value_usd", "loss_given_default"
)]
internal_el_lookup$company_id <- as.character(internal_el_lookup$company_id)
internal_el_lookup <- merge(internal_el_lookup, internal_pd_lookup, by = "company_id")
# Positive magnitude, matching the package-wide EL convention.
internal_el_lookup$internal_el <-
  internal_el_lookup$exposure_value_usd *
  internal_el_lookup$loss_given_default *
  internal_el_lookup$internal_pd

result_el <- integrate_el(
  analysis_data,
  internal_el = internal_el_lookup[, c("company_id", "internal_el")]
)
# default method is "zscore": effective-PD probit recombination, zero-safe.
```

## Interpretation

### EL sign convention, LGD, EAD, horizon

- **EL sign convention.** All `expected_loss_*` columns and all three
  `integrate_el()` methods store EL as a **positive magnitude**. Direction
  lives in the *adjustment*, not the level: a **positive** EL adjustment means
  **more** expected loss = worse = rendered **red**; a **negative** adjustment
  means less loss = better = **green**. The PD adjustment follows the same
  rule (positive = PD rose = red).
- **EAD.** The `exposure_at_default` column is the Basel **exposure at default**:
  the simple runner sets it to each loan's NPV-share-allocated exposure — the
  gross exposure *before* the LGD haircut (LGD is a fraction *of* EAD). The
  LGD-weighted exposure is a separate column,
  `lgd_weighted_exposure = exposure_at_default * loss_given_default` (= EAD × LGD).
  Every `expected_loss_*` column is `exposure_at_default * loss_given_default * pd`
  (= EAD × LGD × PD). The EL zscore method uses `lgd_weighted_exposure` as its
  denominator when present, otherwise reconstructs it as
  `exposure_value_usd * loss_given_default`.
- **LGD.** `loss_given_default` is the fraction lost on default, in `[0, 1]`, and
  is a fraction *of* EAD. Expected loss is the Basel product
  `EL = EAD × LGD × PD` (= `exposure_at_default * loss_given_default * pd`). The EL
  zscore method back-transforms EL to an effective PD via
  `|EL| / lgd_weighted_exposure` (= `|EL| / (EAD × LGD)`), recombines in probit
  space, then re-scales — so LGD enters through that denominator.
- **Horizon.** TRISK PDs here are **term-structured** to the loan `term`
  (multi-year), not a 12-month point-in-time PD. Treat the integrated PD/EL as
  a **lifetime / multi-year** quantity, closer in spirit to an IFRS-9 Stage 2/3
  lifetime ECL input than to a 12-month regulatory PD. Do not feed it directly
  into a 12-month-PD slot without re-annualising.

### The seven pipeline artefacts

The pipeline ships seven ready-made plots/tables. Each has a one-line
"when to use".

**N1 — PD method comparison.** The three methods overlaid: a black tick for
the internal PD, a coloured shape per method for its adjusted PD. Wide spread
means the choice of method moves the number a lot.
*When to use: the methodology choice is on the table (committee review,
regulator pushback). Tight clustering means picking a method is a non-decision.*

```{r fig.width=7, fig.height=4}
pipeline_trisk_pd_method_comparison(analysis_data,
                                     internal_pd = internal_pd_lookup)
```

On sparse portfolios, `granularity = "firm"` and `scale = "pseudo_log"` open
up the per-firm picture when sector aggregation or Merton underflow collapses
everything against the axis:

```{r fig.width=8, fig.height=5}
pipeline_trisk_pd_method_comparison(
  analysis_data,
  internal_pd = internal_pd_lookup,
  granularity = "firm",
  scale       = "pseudo_log"
)
```

**N2 — PD waterfall.** Per-sector decomposition: Internal -> signed Adjustment
-> Adjusted. The middle bar flips fill on sign (red worsens, green improves),
so the "before, change, after" reads at a glance.
*When to use: non-quantitative audiences read waterfalls faster than grouped
bars; the message is "how much does integration move PD".*

```{r fig.width=7, fig.height=4}
pipeline_trisk_pd_waterfall(result_zs)
```

**P1 — PD integration bars.** Four bars per group: Internal PD, TRISK Baseline,
TRISK Shock, TRISK-Adjusted PD. The gap between Internal and Adjusted is the
integration effect; Baseline and Shock show where the model itself moved.
*When to use: first plot to show a counterparty — "what happens to my PDs if I
integrate TRISK?" Supports `granularity = "firm"` / `scale = "pseudo_log"`.*

```{r fig.width=7, fig.height=4}
pipeline_trisk_pd_integration_bars(result_zs)
```

```{r fig.width=8, fig.height=6}
pipeline_trisk_pd_integration_bars(
  result_zs,
  granularity = "firm",
  scale       = "pseudo_log"
)
```

**P2 — EL adjustment bars.** Horizontal bars of the signed EL delta
(Adjusted minus Internal) by sector. Red = integration says expect **more**
loss than your own model; green = less. Levels are not on this plot — only the
adjustment.
*When to use: the question is sign and magnitude of EL impact by sector —
"where does TRISK think you're under-reserving?"*

```{r fig.width=7, fig.height=4}
pipeline_trisk_el_adjustment_bars(result_el)
```

**P3a — PD KPI table.** One-row portfolio summary: weighted internal PD,
weighted adjusted PD, the adjustment in pp, and direction.
*When to use: top of a report or first slide; portfolio-level headline only.*

```{r}
pipeline_trisk_pd_kpi_table(result_zs$aggregate)
```

**P3b — EL KPI table.** One-row EL summary. The headline bps metric is the
**adjusted EL level as a loss rate**: `Adjusted EL / exposure` in bps
(`el_adjusted_bps`) — the total expected-loss rate of the shocked book. The
**climate overlay (delta)**, `EL adjustment / exposure` (`el_adjustment_bps`), is
shown as a secondary column: it isolates the marginal effect of the transition
scenario. Both use *notional* exposure (EAD) as the denominator: dividing by
`lgd_weighted_exposure` (= EAD × LGD) would instead yield PD-in-bps, not a loss rate.
*When to use: same as P3a but for EL; lead with the adjusted EL level, then point
to the delta to attribute how much of it is climate.*

```{r}
pipeline_trisk_el_kpi_table(result_el$aggregate)
```

**P4 — EL sector breakdown.** One row per sector: exposure, internal EL,
adjusted EL, signed delta, direction arrow, and the adjusted EL level as a loss
rate (`Adjusted EL / exposure`, bps). Sits between P2 (deltas only) and P3b (one
number).
*When to use: the reference table for any EL discussion; the bps column makes
sectors with different exposure magnitudes comparable.*

```{r}
pipeline_trisk_el_sector_breakdown_table(result_el$portfolio)
```

**Reading order for a bank-facing report:** P3a + P3b (headline), then N2 + P2
(the change story), then P1 + P4 (full reference levels), and N1 only if the
method choice is being questioned.

### Custom views

The pipeline plots cover the standard story; a couple of custom chunks round it
out. Expected loss by sector, baseline vs shock:

```{r fig.width=8, fig.height=4}
el_by_sector <- analysis_data %>%
  dplyr::group_by(sector) %>%
  dplyr::summarise(
    baseline = sum(.data$expected_loss_baseline, na.rm = TRUE),
    shock    = sum(.data$expected_loss_shock,    na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  tidyr::pivot_longer(
    cols = c("baseline", "shock"), names_to = "scenario", values_to = "el"
  )

ggplot2::ggplot(el_by_sector,
                ggplot2::aes(x = sector, y = .data$el, fill = .data$scenario)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8),
                    width = 0.7) +
  ggplot2::scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
  ggplot2::scale_fill_manual(values = c(baseline = "#5D9324", shock = "#F53D3F")) +
  TRISK_PLOT_THEME_FUNC() +
  ggplot2::labs(x = "Sector", y = "Expected loss (USD, millions)",
                fill = "Scenario",
                title = "Expected loss by sector: baseline vs shock")
```

Internal EL vs TRISK-Adjusted EL, per sector (both are positive magnitudes):

```{r fig.width=8, fig.height=4}
el_compare_sector <- result_el$portfolio %>%
  dplyr::group_by(sector) %>%
  dplyr::summarise(
    `Internal EL`        = sum(.data$internal_el, na.rm = TRUE),
    `TRISK-Adjusted EL`  = sum(.data$trisk_adjusted_el, na.rm = TRUE),
    .groups              = "drop"
  ) %>%
  tidyr::pivot_longer(
    cols = c("Internal EL", "TRISK-Adjusted EL"),
    names_to = "el_type", values_to = "el"
  )

ggplot2::ggplot(el_compare_sector,
                ggplot2::aes(x = sector, y = .data$el, fill = .data$el_type)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8),
                    width = 0.7) +
  ggplot2::scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
  ggplot2::scale_fill_manual(values = c(`Internal EL` = "#BAB6B5",
                                        `TRISK-Adjusted EL` = "#AA2A2B")) +
  TRISK_PLOT_THEME_FUNC() +
  ggplot2::labs(x = "Sector", y = "Expected loss (USD, millions)",
                fill = "",
                title = "Internal EL vs TRISK-Adjusted EL by sector")
```

## Caveats / Limitations

- **PD level is not portable; only the shift is.** TRISK's Merton PD level is
  calibration-specific. The integration deliberately transports the
  baseline-to-shock *change*, not TRISK's absolute PD.
- **Zero-baseline rows lose signal under `relative`.** When `pd_baseline = 0`
  the relative method returns the internal PD unchanged. Use `absolute` or
  `zscore` if that matters for your book.
- **Clipping biases the extreme tails.** `zscore_floor` / `zscore_cap` bound
  PDs before `qnorm()`. Values that should sit below the floor (or above the
  cap) are pulled in, so the recombination is approximate at the extremes.
- **Z-score is probit recombination, not Basel IRB.** It does not apply the
  Basel asset-correlation / Vasicek conditional-PD capital formula. There is no
  correlation parameter and no supervisory confidence level. Do not present its
  output as a regulatory capital number.
- **Horizon mismatch.** Integrated PD/EL is term-structured (multi-year),
  closer to lifetime ECL than to a 12-month regulatory PD. Re-annualise before
  using it in a 12-month slot — `pd_lifetime_to_annual(pd, term)` does this under
  a constant-hazard assumption (and `pd_annual_to_lifetime()` is its inverse).
- **Static LGD (L1).** The transition shock moves PD and NPV, but
  `loss_given_default` is held constant — there is no downturn/stress LGD. EL
  changes here reflect PD/value effects only; if your framework uses
  scenario-conditional or downturn LGD, apply it separately.
- **Geography filtering drops out-of-region assets (G1).** With
  `scenario_geography` other than `"Global"`, the model keeps only assets in the
  selected region(s); out-of-region assets are excluded from the run silently.
  Reconcile your input asset count against the results if you scope by geography.
- **EL zscore normalizer basis must match.** The effective-PD round-trip is only
  correct when the EL columns and the EAD×LGD denominator are on the same basis.
  When EAD is NPV-share-allocated (as the simple runner does), supply the matching
  `lgd_weighted_exposure` column rather than relying on the
  `exposure_value_usd * loss_given_default` fallback.
- **`internal_pd` must be in `[0, 1]` and joinable by `company_id`.** Unmatched
  `company_id`s fall back to `pd_baseline` with a warning; check the warning
  before trusting the result.

## See also

- `getting-started` — first run of TRISK end to end.
- `inputs-and-outputs` — the input files and output schema in detail.
- `simple-portfolio-analysis` — building and running a portfolio.
- `sensitivity-analysis` — varying scenarios and parameters.
