---
title: "Bank workflow 2: Simple portfolio analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bank workflow 2: Simple portfolio analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(dplyr)
library(magrittr)
```

## Overview

This vignette walks through running TRISK on a bank loan book with
`run_trisk_on_simple_portfolio()` — the bank entry point. You bring company
exposures, terms, and loss-given-default (no `country_iso2` column needed), and
TRISK runs and allocates each loan's exposure across the company's technologies
for you. The per-technology results feed the equity- (NPV, exposure) and
credit-risk (PD, expected-loss) plots directly.

## Inputs

> **Input data — where your data goes.** TRISK needs **five inputs**: four that
> describe the world — **assets**, **scenarios**, **NGFS carbon price** and
> **financial features** — plus your **portfolio**. The main portfolio file is
> **`portfolio_ids`** (matched by `company_id`); `portfolio_names` and
> `portfolio_countries` are options. **The CSVs loaded below are placeholders**
> (bundled samples) — replace them with your own files. See
> [Inputs and outputs](bank_1_inputs-and-outputs.html) for `setup_trisk_inputs()`
> and the `trisk_inputs/` folder convention.

The simple runner needs the four TRISK model inputs, shipped as test data in the
`trisk.model` package:

```{r}
assets_testdata <- read.csv(system.file("testdata", "assets_testdata.csv", package = "trisk.model", mustWork = TRUE))
# NGFS 2024 scenarios start in 2023; the bundled assets reach back to 2022 and
# TRISK errors on assets outside the scenario window, so scope to 2023 onward.
assets_testdata <- assets_testdata[assets_testdata$production_year >= 2023, ]
scenarios_testdata <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model", mustWork = TRUE))
financial_features_testdata <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model", mustWork = TRUE))
ngfs_carbon_price_testdata <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model", mustWork = TRUE))
```

They also share the same scenario settings — a baseline and a target scenario,
plus the geography to evaluate:

```{r}
baseline_scenario <- "NGFS2024GCAM_CP"
target_scenario <- "NGFS2024GCAM_NZ2050"
scenario_geography <- "Global"
```

The portfolio schema is covered in the example below.

## Minimal example

`run_trisk_on_simple_portfolio()` expects a portfolio with just five columns:

- `company_id`
- `company_name`
- `exposure_value_usd`
- `term`
- `loss_given_default`

No `country_iso2` is required. Load the bundled simple portfolio:

```{r}
simple_portfolio <- read.csv(
  system.file("testdata", "simple_portfolio.csv", package = "trisk.analysis", mustWork = TRUE)
)
simple_portfolio
```

Run the model:

```{r}
simple_results <- run_trisk_on_simple_portfolio(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  portfolio_data = simple_portfolio,
  baseline_scenario = baseline_scenario,
  target_scenario = target_scenario,
  scenario_geography = scenario_geography
)

portfolio_results_tech_detail <- simple_results$portfolio_results_tech_detail
portfolio_results <- simple_results$portfolio_results
```

### NPV-based exposure allocation

The simple runner adds a column the full runner does not: `exposure_value_usd_share`.
A single loan exposure has to be spread across the company's technologies before
technology-level risk can be attributed to it. TRISK allocates exposure from
baseline NPV shares at company/sector/technology level:

1. compute baseline NPV share per run;
2. allocate exposure with that share;
3. average after dropping `run_id`;
4. re-scale so exposure totals match the original portfolio exposure.

```{r}
portfolio_results_tech_detail |>
  dplyr::select(
    company_id, term, sector, technology,
    exposure_value_usd_share,
    net_present_value_baseline
  ) |>
  utils::head(10)
```

Because `exposure_value_usd_share` is computed after dropping `run_id`, it is
constant across runs for a given `(company_id, term, sector, technology)`.

```{r}
exposure_share_check <- portfolio_results_tech_detail |>
  dplyr::distinct(
    company_id, term, sector, technology,
    exposure_value_usd_share
  ) |>
  dplyr::group_by(company_id, term) |>
  dplyr::summarise(allocated_exposure = sum(exposure_value_usd_share, na.rm = TRUE), .groups = "drop") |>
  dplyr::left_join(
    portfolio_results |>
      dplyr::group_by(company_id, term) |>
      dplyr::summarise(original_exposure = sum(exposure_value_usd, na.rm = TRUE), .groups = "drop"),
    by = c("company_id", "term")
  ) |>
  dplyr::mutate(gap = allocated_exposure - original_exposure)

exposure_share_check
```

The `gap` column should be close to zero (floating-point tolerance) — confirming
allocated exposure reconciles back to the original loan amounts.

## Plotting the per-technology results

The simple runner's `portfolio_results_tech_detail` carries per-technology NPV,
PD, and expected-loss columns — exactly what the package plotting helpers need.
Expose the NPV-share-allocated exposure (`exposure_at_default`) as
`exposure_value_usd` so the plots can aggregate it, then pass the frame in:

```{r}
npv_analysis <- portfolio_results_tech_detail |>
  dplyr::mutate(exposure_value_usd = exposure_at_default)
```

**Equity risk** — average percentage NPV change per technology:

```{r}
pipeline_trisk_npv_change_plot(npv_analysis)
```

Resulting portfolio exposure change:

```{r}
pipeline_trisk_exposure_change_plot(npv_analysis)
```

**Bonds & loans risk** — average PDs at baseline and shock:

```{r}
pipeline_trisk_pd_term_plot(npv_analysis)
```

Resulting portfolio expected loss:

```{r}
pipeline_trisk_expected_loss_plot(npv_analysis)
```

## Technology mix: NPV value vs exposure

`pipeline_trisk_tech_mix()` shows each sector's technology composition two ways —
as a share of baseline NPV value and as a share of exposure (EAD) — with
technologies rendered as shades of the sector's colour (darkest = most
carbon-intensive). The two bases can disagree: a technology can be a large share
of *value* yet a small share of *exposure* (or the reverse), so reporting only
one basis can mislead the credit read. On a small bundled book a sector may
resolve to a single technology; richer multi-technology counterparties fill the
mix out.

```{r fig.width=8, fig.height=5}
pipeline_trisk_tech_mix(npv_analysis)
```

## Interpretation

- The **simple runner** is the fastest way to attribute climate transition risk
  to a known book. Its `exposure_value_usd_share` lets you see how each loan's
  exposure spreads across a company's technologies, so you can spot which
  technology mix drives a counterparty's risk.
- A negative NPV change and a rising shock PD on the same technology flag where
  the transition both erodes asset value and lifts default probability — the
  positions to scrutinise first.

## Caveats

- Exposure allocation in the simple runner is NPV-share based and reconciles to
  the original totals only up to floating-point tolerance — confirm the `gap`
  before relying on allocated figures.
- The scenarios, geography, and bundled test data here are illustrative. Swap in
  your own asset, scenario, financial, and carbon-price inputs for production
  analysis.

## See also

- `getting-started` — install and run a first TRISK analysis.
- `inputs-and-outputs` — the input schemas and output tables in detail.
- `pd-el-integration` — turning TRISK output into PD and expected-loss measures.
- `sensitivity-analysis` — varying scenarios and parameters across runs.
