---
title: "Bank workflow 1: Bank credit risk analysis — start here"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bank workflow 1: Bank credit risk analysis — start here}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(dplyr)
library(magrittr)
```

This is the **entry point for a bank** running TRISK on its own loan book. It
shows the five inputs TRISK needs, where your data goes, and a first end-to-end
run. The vignettes that follow ([Inputs and outputs](bank_2_inputs-and-outputs.html),
[Run on a portfolio](bank_3_run-on-a-portfolio.html),
[Sensitivity analysis](bank_4_sensitivity-analysis.html),
[PD & EL integration](bank_5_pd-el-integration.html)) go deeper; start here.

## The five inputs

TRISK takes **five inputs**. The first four describe the world; the fifth is
**your portfolio**.

| Input | Argument | What it is | Who provides it |
|-------|----------|-----------|-----------------|
| Assets | `assets_data` | Physical/production assets per company | **You** |
| Scenarios | `scenarios_data` | Climate-scenario price & production pathways | Download (full set) / sample bundled |
| NGFS carbon price | `carbon_data` | Carbon-price trajectories | Sample bundled |
| Financial features | `financial_data` | Per-company financials (PD, margins, leverage, volatility) | **You** |
| **Portfolio** | `portfolio_data` | **Your loan book — `portfolio_ids.csv`** | **You** |

## Set up your input folder

Run this **once** to create a `trisk_inputs/` folder in your working directory,
pre-filled with blank templates and filled samples plus a README:

```{r eval=FALSE}
setup_trisk_inputs()
```

It creates:

```
trisk_inputs/
  README.md           # required files, column schemas, portfolio modes
  templates/          # blank, header-only CSVs to fill in
    assets.csv  scenarios.csv  ngfs_carbon_price.csv
    financial_features.csv  portfolio_ids.csv
  samples/            # the bundled example data, as a worked reference
    (same five files, filled)
```

> The folder lives in **your project directory**, never inside the installed
> package — the package library is read-only on many systems and is overwritten
> on every update, so your data must live in your own project.

Fill the templates (or copy a `samples/` file and edit it), save each as
`trisk_inputs/<input>.csv`, then point the reads below at those files. For the
full production scenario library, run
`download_trisk_inputs(local_save_folder = "trisk_inputs")` — it currently
fetches `scenarios.csv`; you supply `assets.csv`, `financial_features.csv` and
`ngfs_carbon_price.csv` yourself, starting from the bundled samples.

## Your portfolio is the centre: `portfolio_ids`

`portfolio_ids.csv` is the **main** portfolio input. One row per loan, keyed by
`company_id`:

```{r}
portfolio_ids_testdata <- read.csv(
  system.file("testdata", "portfolio_ids_testdata.csv", package = "trisk.analysis")
)
```

```{r echo=FALSE}
knitr::kable(portfolio_ids_testdata) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "260px")
```

Matching companies by `company_id` is the recommended path. Two alternatives are
available as **options**:

- **`portfolio_names.csv`** — same columns, but companies are fuzzy-matched by
  `company_name` when you do not have `company_id`.
- **`portfolio_countries.csv`** — match by `country_iso2` only (country-level
  aggregation) when neither id nor name is available.

For the [Sensitivity analysis](bank_4_sensitivity-analysis.html) and
[PD & EL integration](bank_5_pd-el-integration.html) workflows, add an `internal_pd`
column (your own PD per company, in `[0, 1]`) — that is the
`portfolio_ids_internal_pd` variant those vignettes use.

## A first run on placeholder data

The CSVs loaded here are **placeholders** (sample data bundled with the
packages). Replace them with your own files in `trisk_inputs/` once you have
filled the templates — the commented block shows the swap.

```{r}
# --- PLACEHOLDERS: sample data shipped with the packages ---
assets_data    <- read.csv(system.file("testdata", "assets_testdata.csv",             package = "trisk.model"))
scenarios_data <- read.csv(system.file("testdata", "scenarios_testdata.csv",          package = "trisk.model"))
carbon_data    <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv",  package = "trisk.model"))
financial_data <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model"))
portfolio_data <- read.csv(system.file("testdata", "portfolio_ids_testdata.csv",      package = "trisk.analysis")) # MAIN portfolio input

# --- PRODUCTION: replace the placeholders above with your own data ---
# assets_data    <- read.csv("trisk_inputs/assets.csv")
# scenarios_data <- read.csv("trisk_inputs/scenarios.csv")
# carbon_data    <- read.csv("trisk_inputs/ngfs_carbon_price.csv")
# financial_data <- read.csv("trisk_inputs/financial_features.csv")
# portfolio_data <- read.csv("trisk_inputs/portfolio_ids.csv")   # or portfolio_names.csv / portfolio_countries.csv
```

Pick a baseline and a target scenario (from the same provider), then run:

```{r}
analysis_data <- run_trisk_on_portfolio(
  assets_data        = assets_data,
  scenarios_data     = scenarios_data,
  financial_data     = financial_data,
  carbon_data        = carbon_data,
  portfolio_data     = portfolio_data,
  baseline_scenario  = "NGFS2023GCAM_CP",
  target_scenario    = "NGFS2023GCAM_NZ2050",
  scenario_geography = "Global"
)
```

```{r echo=FALSE}
knitr::kable(head(analysis_data[, c("company_id", "sector", "technology", "term",
                                    "pd_baseline", "pd_shock",
                                    "net_present_value_baseline", "net_present_value_shock")])) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "260px")
```

That is a full TRISK run on your portfolio. From here:

- **[Inputs and outputs](bank_2_inputs-and-outputs.html)** — every input column and
  every output table, in detail.
- **[Run on a portfolio](bank_3_run-on-a-portfolio.html)** — the simple vs full runners,
  exposure allocation, and the plotting helpers.
- **[Sensitivity analysis](bank_4_sensitivity-analysis.html)** — sweep shock year, IAM
  and ambition, read through a bank-impact lens.
- **[PD & EL integration](bank_5_pd-el-integration.html)** — recombine TRISK PDs with
  your internal PDs and translate to expected-loss basis points.
