---
title: "inputs-and-outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inputs-and-outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(magrittr)
```


# Input datasets

Load the internal datasets

```{r}
assets_testdata <- read.csv(system.file("testdata", "assets_testdata.csv", package = "trisk.model"))
scenarios_testdata <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model"))
financial_features_testdata <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model"))
ngfs_carbon_price_testdata <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model"))
```

### Assets Test Data

This dataset contains data about company assets, including production, technology, and geographical details.

#### Data Description

The `assets_testdata` dataset includes the following columns:

- `company_id`: Unique identifier for the company.
- `company_name`: Name of the company.
- `asset_id`: Unique identifier for the asset.
- `country_iso2`: ISO 3166-1 alpha-2 code for the country.
- `asset_name`: Name of the asset.
- `production_year`: Year of production data.
- `emission_factor`: Emissions from production.
- `technology`: Type of technology used.
- `sector`: Production sector.
- `capacity`: Asset capacity.
- `capacity_factor`: Asset utilization percentage.
- `production_unit`: Unit for production.

#### Data Structure

```{r}
str(assets_testdata)
```

#### Sample Data

```{r}
knitr::kable(head(assets_testdata)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px") %>%
  kableExtra::column_spec(1:ncol(assets_testdata), width = "150px")
```

---

### Financial Features Test Data

This dataset contains financial metrics necessary for calculating stress test outputs.

#### Data Description

The `financial_features_testdata` dataset includes the following columns:

- `company_id`: Unique identifier for the company.
- `pd`: Probability of default for the company.
- `net_profit_margin`: Net profit margin for the company.
- `debt_equity_ratio`: Debt to equity ratio.
- `volatility`: Volatility of the company's asset values.

#### Data Structure

```{r}
str(financial_features_testdata)
```

#### Sample Data

```{r}
knitr::kable(head(financial_features_testdata)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

---

### NGFS Carbon Price Test Data

This dataset provides carbon pricing data used in the stress test scenarios.

#### Data Description

The `ngfs_carbon_price_testdata` dataset includes the following columns:

- `year`: Year of the carbon price.
- `model`: Model used to generate the carbon price.
- `scenario`: Scenario name.
- `scenario_geography`: Geographic region for the scenario.
- `variable`: The variable measured (e.g., carbon price).
- `unit`: Unit of the variable.
- `carbon_tax`: The amount of carbon tax applied in the scenario.

#### Data Structure

```{r}
str(ngfs_carbon_price_testdata)
```

#### Sample Data

```{r}
knitr::kable(head(ngfs_carbon_price_testdata)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

---

### Scenarios Test Data

This dataset contains scenario-specific data including price paths, capacity factors, and other relevant information.

#### Data Description

The `scenarios_testdata` dataset includes the following columns:

- `scenario_geography`: Region relevant to the scenario.
- `scenario`: Scenario name.
- `scenario_pathway`: Specific pathway for the scenario.
- `scenario_type`: Type of scenario (e.g., baseline, shock).
- `sector`: Sector of production.
- `technology`: Type of technology.
- `scenario_year`: Year of the scenario data.
- `scenario_price`: Price in the scenario.
- `price_unit`: Unit for the price.
- `pathway_unit`: Unit of the pathway.
- `technology_type`: Type of technology involved (carbon or renewable).

#### Data Structure

```{r}
str(scenarios_testdata)
```

#### Sample Data


```{r}
knitr::kable(head(scenarios_testdata, 50)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```


# Trisk run and Output datasets  

Define the scenarios to use:
```{r}
baseline_scenario <- "NGFS2023GCAM_CP"
target_scenario <- "NGFS2023GCAM_NZ2050"
scenario_geography <- "Global"
```

This function runs the Trisk model using the provided input and returns the outputs. 

```{r}
st_results <- trisk.model::run_trisk_model(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  baseline_scenario = baseline_scenario,
  target_scenario = target_scenario,
  scenario_geography = scenario_geography
)
```

Get result dataframes from function output

```{r}
npv_results <- st_results$npv_results
pd_results <- st_results$pd_results
company_trajectories <- st_results$company_trajectories
```


### NPV results

#### Data Description

The `npv_results` dataset includes the following columns:

- `run_id`: Unique identifier for the simulation run.
- `company_id`: Unique identifier for the company.
- `asset_id`: Unique identifier for the asset.
- `company_name`: Name of the company.
- `asset_name`: Name of the asset.
- `country_iso2`: ISO 3166-1 alpha-2 code for the country.
- `sector`: Sector in which the company operates (e.g., Oil&Gas, Coal, Power).
- `technology`: Type of technology used by the company (e.g., Gas, CoalCap, RenewablesCap).
- `net_present_value_baseline`: Net present value (NPV) under the baseline scenario.
- `net_present_value_shock`: Net present value (NPV) under the shock scenario.

#### Data Structure

```{r}
str(npv_results)
```

#### Sample Data

```{r}
knitr::kable(head(as.data.frame(npv_results))) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```



### PD results

#### Data Description

The `pd_results` dataset includes the following columns:

- `run_id`: Unique identifier for the simulation run.
- `company_id`: Unique identifier for the company.
- `company_name`: Name of the company.
- `sector`: Sector in which the company operates (e.g., Oil&Gas, Coal).
- `term`: Time period for the probability of default (PD) calculation.
- `pd_baseline`: Probability of default (PD) under the baseline scenario.
- `pd_shock`: Probability of default (PD) under the shock scenario.


#### Data Structure

```{r}
str(pd_results)
```

#### Sample Data

```{r}
knitr::kable(head(as.data.frame(pd_results))) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```



### Company trajectories results

#### Data Description

The `company_trajectories` dataset includes the following columns:

- `run_id`: Unique identifier for the simulation run.
- `asset_id`: Unique identifier for the asset.
- `asset_name`: Name of the asset.
- `company_id`: Unique identifier for the company.
- `company_name`: Name of the company.
- `year`: Year of the scenario data.
- `sector`: Sector in which the company operates (e.g., Oil&Gas, Coal).
- `technology`: Type of technology used by the company.
- `production_plan_company_technology`: Production plan for the company’s technology.
- `production_baseline_scenario`: Production output under the baseline scenario.
- `production_target_scenario`: Production output under the target scenario.
- `production_shock_scenario`: Production output under the shock scenario.
- `pd`: Probability of default for the company.
- `net_profit_margin`: Net profit margin for the company.
- `debt_equity_ratio`: Debt to equity ratio for the company.
- `volatility`: Volatility of the company’s asset values.
- `scenario_price_baseline`: Price under the baseline scenario.
- `price_shock_scenario`: Price under the shock scenario.
- `net_profits_baseline_scenario`: Net profits under the baseline scenario.
- `net_profits_shock_scenario`: Net profits under the shock scenario.
- `discounted_net_profits_baseline_scenario`: Discounted net profits under the baseline scenario.
- `discounted_net_profits_shock_scenario`: Discounted net profits under the shock scenario.

#### Data Structure

```{r}
str(company_trajectories)
```

#### Sample Data

```{r}
knitr::kable(head(as.data.frame(company_trajectories))) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

