---
title: "main-functions-demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{main-functions-demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(trisk.model)
```

# trisk.analysis main functions demo

This vignette demonstrates how to use the `trisk.model` package to download input data, run TRISK models, and perform sensitivity analysis. It walks through the process of running the model, inspecting outputs, and using filters for more detailed analyses.


### Download TRISK Input Data

The TRISK model requires specific input datasets. In this section, we download these datasets from the defined endpoint and store them in a local directory.

```{r}
# Define the endpoint URL and S3 path for downloading data
endpoint_url <- "https://crispy-datamodels-bucket.fra1.cdn.digitaloceanspaces.com"
s3_path <- "crispy-datamodels-bucket/trisk_V2/csv"
local_trisk_inputs_folder <- file.path("..","data-raw", "data", "trisk_inputs")

# Download TRISK input data
download_trisk_inputs(endpoint_url, s3_path, local_trisk_inputs_folder)
```

### Run a Simple TRISK Model

Next, we demonstrate a simple TRISK model run using a baseline and target scenario. We also measure the time taken for the execution.

```{r}
# Set output path for TRISK results
trisk_output_path <- file.path("..","data-raw", "data", "trisk_outputs")

# Measure execution time
start_time <- Sys.time()

# Run TRISK model with baseline and target scenarios
run_trisk(
  input_path = local_trisk_inputs_folder,
  output_path = trisk_output_path,
  baseline_scenario = "Steel_baseline",
  target_scenario = "Steel_NZ",
  scenario_geography = "Global"
)

# End time
end_time <- Sys.time()

# Calculate time taken for execution
time_taken <- end_time - start_time
print(time_taken)
```

### View Latest TRISK Results

After the model has been run, we can inspect the results. TRISK results can include various metrics such as NPV, PD, trajectories, and parameters.

```{r}
# Print the latest TRISK results for different result types
print(get_latest_trisk_result(trisk_output_path = trisk_output_path, result_type = "npv"))
print(get_latest_trisk_result(trisk_output_path = trisk_output_path, result_type = "pd"))
print(get_latest_trisk_result(trisk_output_path = trisk_output_path, result_type = "trajectories"))
print(get_latest_trisk_result(trisk_output_path = trisk_output_path, result_type = "params"))
```

### Run a Filtered TRISK Model (Optional)

You can also run the TRISK model on a filtered subset of assets. This example filters the dataset by country, sector, and technology.

```{r}
# Run the TRISK model with a smaller filtered dataset (currently unused)
single_filtered_results <- run_trisk_single_filtered(
  input_path = local_trisk_inputs_folder,
  run_params = list(
    scenario_geography = "Global",
    baseline_scenario = "Steel_baseline",
    target_scenario = "Steel_NZ",
    shock_year = 2030
  ),
  country_iso2 = c("US", "AR"),
  sector = "Steel",
  technology = c("EAF-DRI", "BOF-BF"),
  company_name = NULL
)

# Print the filtered TRISK results for different result types
print(single_filtered_results["npv"])
print(single_filtered_results["pd"])
print(single_filtered_results["trajectories"])
print(single_filtered_results["params"])
```

### Sensitivity Analysis

Sensitivity analysis allows us to explore how changes in parameters affect the results. In this example, we vary the shock year in multiple TRISK runs.

```{r}
# Define run parameters for sensitivity analysis
run_params <- list(
  list(
    scenario_geography = "Global",
    baseline_scenario = "Steel_baseline",
    target_scenario = "Steel_NZ",
    shock_year = 2030
  ),
  list(
    scenario_geography = "Global",
    baseline_scenario = "Steel_baseline",
    target_scenario = "Steel_NZ",
    shock_year = 2025
  )
)

# Run sensitivity analysis across multiple TRISK runs
sensitivity_analysis_results <- run_trisk_sa(
  input_path = local_trisk_inputs_folder,
  run_params = run_params
)

# Print sensitivity analysis results for different result types
print(sensitivity_analysis_results["npv"])
print(sensitivity_analysis_results["pd"])
print(sensitivity_analysis_results["trajectories"])
print(sensitivity_analysis_results["params"])
```

### Sensitivity Analysis on Filtered Subset

You can also perform sensitivity analysis on a filtered subset of assets. This example filters the dataset by country, sector, and technology, and then runs the analysis.

```{r}
# Run sensitivity analysis on a subset of assets
sensitivity_analysis_results_on_filtered_assets <- run_trisk_sa(
  input_path = local_trisk_inputs_folder,
  run_params = run_params,
  country_iso2 = c("US", "AR"),
  sector = "Steel",
  technology = c("EAF-DRI", "BOF-BF")
)

# Print sensitivity analysis results for the filtered subset
print(sensitivity_analysis_results_on_filtered_assets["npv"])
print(sensitivity_analysis_results_on_filtered_assets["pd"])
print(sensitivity_analysis_results_on_filtered_assets["trajectories"])
print(sensitivity_analysis_results_on_filtered_assets["params"])
```
