---
title: "pd-el-integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pd-el-integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(magrittr)
```

# PD and EL Integration

## 1. Setup

Load the bundled testdata shipped with `trisk.model` and `trisk.analysis`.
The portfolio file is `portfolio_ids_internal_pd_testdata.csv`, which
augments the classic `portfolio_ids_testdata.csv` with an `internal_pd`
column per exposure. That column is **mandatory** — integration is
meaningless without the bank's own PD to integrate TRISK's shift into, so
the setup fails loudly if it's missing.

```{r}
assets_testdata    <- read.csv(system.file("testdata", "assets_testdata.csv",    package = "trisk.model"))
scenarios_testdata <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model"))
fin_testdata       <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model"))
carbon_testdata    <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model"))
portfolio_ids_internal_pd <- read.csv(system.file("testdata", "portfolio_ids_internal_pd_testdata.csv",
                                                  package = "trisk.analysis"))

stopifnot(
  "Portfolio file must include an `internal_pd` column per exposure." =
    "internal_pd" %in% colnames(portfolio_ids_internal_pd),
  "`internal_pd` values must be numeric in [0, 1]." =
    is.numeric(portfolio_ids_internal_pd$internal_pd) &&
      all(portfolio_ids_internal_pd$internal_pd >= 0 &
          portfolio_ids_internal_pd$internal_pd <= 1, na.rm = TRUE)
)
```

## 2. Run TRISK on the portfolio

The `internal_pd` column is carried alongside — TRISK only needs the standard
portfolio schema, and we keep the internal PDs as a separate lookup table to
feed into `integrate_pd()` later.

```{r}
analysis_data <- run_trisk_on_portfolio(
  assets_data       = assets_testdata,
  scenarios_data    = scenarios_testdata,
  financial_data    = fin_testdata,
  carbon_data       = carbon_testdata,
  portfolio_data    = portfolio_ids_internal_pd,
  baseline_scenario = "NGFS2023GCAM_CP",
  target_scenario   = "NGFS2023GCAM_NZ2050"
)

# The internal_pd lookup used throughout the rest of the vignette:
internal_pd_lookup <- portfolio_ids_internal_pd[, c("company_id", "internal_pd")]
```

```{r echo=FALSE}
knitr::kable(head(analysis_data[, c("company_id", "sector", "technology", "term",
                                    "pd_baseline", "pd_shock",
                                    "net_present_value_baseline", "net_present_value_shock")])) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "300px")
```

## 3. Bank-risk context (before integration)

Before layering the integration methods on top, look at the raw bank-risk
picture that TRISK produces. The four standard portfolio plots answer "what
did the shock do?" at the equity (NPV, exposure) and credit (PD term, EL)
levels, using the same `analysis_data`.

```{r fig.width=7, fig.height=4}
pipeline_crispy_pd_term_plot(analysis_data)
```

Expected loss by sector (baseline vs shock), all sectors on a single row:

```{r fig.width=8, fig.height=4}
ad_el_preview <- compute_analysis_metrics(analysis_data)
el_by_sector <- ad_el_preview %>%
  dplyr::group_by(sector) %>%
  dplyr::summarise(
    baseline = sum(.data$expected_loss_baseline, na.rm = TRUE),
    shock    = sum(.data$expected_loss_shock,    na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  tidyr::pivot_longer(
    cols = c("baseline", "shock"), names_to = "scenario", values_to = "el"
  )

ggplot2::ggplot(el_by_sector,
                ggplot2::aes(x = sector, y = .data$el, fill = .data$scenario)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8),
                    width = 0.7) +
  ggplot2::scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
  ggplot2::scale_fill_manual(values = c(baseline = "#5D9324", shock = "#F53D3F")) +
  TRISK_PLOT_THEME_FUNC() +
  ggplot2::labs(x = "Sector", y = "Expected loss (USD, millions)",
                fill = "Scenario",
                title = "Expected loss by sector: baseline vs shock")
```

## 4. Why integration matters

TRISK recomputes PD from a Merton structural credit model. That PD level is not
directly comparable to the internal PD your institution already uses; only the
change from baseline to shock carries meaning. Three integration methods
translate the TRISK shift into your internal PD scale.

## 5. Method 1 - Absolute

```{r}
result_abs <- integrate_pd(analysis_data,
                           internal_pd = internal_pd_lookup,
                           method      = "absolute")
```

```{r echo=FALSE}
knitr::kable(head(result_abs$portfolio[, c("company_id", "sector", "internal_pd",
                                           "pd_baseline", "pd_shock",
                                           "trisk_adjusted_pd", "pd_adjustment")])) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "300px")
```

## 6. Method 2 - Relative

> **Note:** when `pd_baseline = 0` the relative method returns the internal PD
> unchanged; the shock signal is lost on zero-baseline rows. Use `absolute` or
> `zscore` if this matters.

```{r}
result_rel <- integrate_pd(analysis_data,
                           internal_pd = internal_pd_lookup,
                           method      = "relative")
```

## 7. Method 3 - Z-score (Basel IRB)

The z-score method combines PDs in the normal-quantile space, preserving the
non-linear relationship at distribution tails. Recommended for Basel IRB-aligned
institutions.

```{r}
result_zs <- integrate_pd(analysis_data,
                          internal_pd = internal_pd_lookup,
                          method      = "zscore")
```

## 8. Overriding internal PDs on the fly

If you want to stress-test with a hypothetical flat internal PD (for example,
as a sanity check), build a vector or alternate lookup and pass it instead:

```{r}
flat_internal <- rep(0.03, nrow(analysis_data))
result_custom <- integrate_pd(analysis_data,
                              internal_pd = flat_internal,
                              method      = "zscore")
```

## 9. Method comparison

```{r fig.width=7, fig.height=4}
pipeline_crispy_pd_method_comparison(analysis_data,
                                     internal_pd = internal_pd_lookup)
```

### Advanced: firm-level and pseudo-log scale

On sparse portfolios where many firms have baseline PDs near zero (Merton
underflow) and a few have visible signal, the default sector-aggregate linear
view can bunch everything against zero. Two arguments open up the per-firm
picture:

- `granularity = "firm"` — one row per firm instead of an EAD-weighted
  sector aggregate. Reveals within-sector method divergence.
- `scale = "pseudo_log"` — x-axis uses `scales::pseudo_log_trans(sigma = 1e-5)`.
  Zero-safe and spreads values across many orders of magnitude.

```{r fig.width=8, fig.height=5}
pipeline_crispy_pd_method_comparison(
  analysis_data,
  internal_pd = internal_pd_lookup,
  granularity = "firm",
  scale       = "pseudo_log"
)
```

## 10. Integration charts

```{r fig.width=7, fig.height=4}
pipeline_crispy_pd_integration_bars(result_zs)
```

The same `granularity` + `scale` arguments work on the bar plot:

```{r fig.width=8, fig.height=6}
pipeline_crispy_pd_integration_bars(
  result_zs,
  granularity = "firm",
  scale       = "pseudo_log"
)
```

## 11. EL integration

`integrate_el()` expects `expected_loss_baseline` and `expected_loss_shock`
columns, which `run_trisk_on_portfolio()` does not produce on its own. Call
`compute_analysis_metrics()` first to derive them as EAD * PD.

For the bank's internal EL, we derive it the same way: `EAD * LGD * internal_pd`.

```{r}
analysis_data_el <- compute_analysis_metrics(analysis_data)
internal_el_lookup <- merge(
  analysis_data_el[, c("company_id", "exposure_value_usd", "loss_given_default")],
  internal_pd_lookup,
  by = "company_id"
)
internal_el_lookup$internal_el <-
  -internal_el_lookup$exposure_value_usd *
   internal_el_lookup$loss_given_default *
   internal_el_lookup$internal_pd

result_el <- integrate_el(analysis_data_el,
                          internal_el = internal_el_lookup[, c("company_id", "internal_el")])
# default method is "zscore" (Basel IRB-aligned, zero-safe via clipping)
```

```{r fig.width=7, fig.height=4}
pipeline_crispy_el_adjustment_bars(result_el)
```

### Internal EL vs TRISK-Adjusted EL

Side-by-side comparison so the integration effect is visible per group. EL is
negative by package convention (loss as negative number); `abs()` is applied
for display so the bars read as magnitudes of expected loss.

**Per sector** — all sectors on one row:

```{r fig.width=8, fig.height=4}
el_compare_sector <- result_el$portfolio %>%
  dplyr::group_by(sector) %>%
  dplyr::summarise(
    `Internal EL`        = sum(.data$internal_el, na.rm = TRUE),
    `TRISK-Adjusted EL`  = sum(.data$trisk_adjusted_el, na.rm = TRUE),
    .groups              = "drop"
  ) %>%
  tidyr::pivot_longer(
    cols = c("Internal EL", "TRISK-Adjusted EL"),
    names_to = "el_type", values_to = "el"
  ) %>%
  dplyr::mutate(el_abs = abs(.data$el))

ggplot2::ggplot(el_compare_sector,
                ggplot2::aes(x = sector, y = .data$el_abs, fill = .data$el_type)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8),
                    width = 0.7) +
  ggplot2::scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
  ggplot2::scale_fill_manual(values = c(`Internal EL` = "#BAB6B5",
                                        `TRISK-Adjusted EL` = "#AA2A2B")) +
  TRISK_PLOT_THEME_FUNC() +
  ggplot2::labs(x = "Sector", y = "|Expected loss| (USD, millions)",
                fill = "",
                title = "Internal EL vs TRISK-Adjusted EL by sector")
```

**Per exposure** — one group of two bars per firm/technology:

```{r fig.width=11, fig.height=5}
el_compare_firm <- result_el$portfolio %>%
  dplyr::mutate(firm = paste(.data$company_id, .data$technology, sep = "/")) %>%
  dplyr::select("firm", "sector",
                `Internal EL` = "internal_el",
                `TRISK-Adjusted EL` = "trisk_adjusted_el") %>%
  tidyr::pivot_longer(
    cols = c("Internal EL", "TRISK-Adjusted EL"),
    names_to = "el_type", values_to = "el"
  ) %>%
  dplyr::mutate(el_abs = abs(.data$el))

ggplot2::ggplot(el_compare_firm,
                ggplot2::aes(x = .data$firm, y = .data$el_abs, fill = .data$el_type)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8),
                    width = 0.7) +
  ggplot2::scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
  ggplot2::scale_fill_manual(values = c(`Internal EL` = "#BAB6B5",
                                        `TRISK-Adjusted EL` = "#AA2A2B")) +
  ggplot2::facet_grid(. ~ sector, scales = "free_x", space = "free_x") +
  TRISK_PLOT_THEME_FUNC() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::labs(x = "Firm / technology", y = "|Expected loss| (USD, millions)",
                fill = "",
                title = "Internal EL vs TRISK-Adjusted EL by exposure")
```

## 12. Portfolio-level KPIs

```{r}
pipeline_crispy_pd_kpi_table(result_zs$aggregate)
```

```{r}
pipeline_crispy_el_kpi_table(result_el$aggregate)
```

## 13. Sector breakdown

```{r}
pipeline_crispy_el_sector_breakdown_table(result_el$portfolio)
```
