---
title: "portfolio-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{portfolio-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(magrittr)
```

# Restrict the analysis to a portfolio

## Generate outputs
### Load the test data

Load the internal datasets

```{r}
assets_testdata <- read.csv(system.file("testdata", "assets_testdata.csv", package = "trisk.model", mustWork = TRUE))
scenarios_testdata <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model", mustWork = TRUE))
financial_features_testdata <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model", mustWork = TRUE))
ngfs_carbon_price_testdata <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model", mustWork = TRUE))
```


### Prepare portfolio

There are 3 possible portfolio input structures :

```{r}
portfolio_countries_testdata <- read.csv(system.file("testdata", "portfolio_countries_testdata.csv", package = "trisk.analysis"))
portfolio_ids_testdata <- read.csv(system.file("testdata", "portfolio_ids_testdata.csv", package = "trisk.analysis"))
portfolio_names_testdata <- read.csv(system.file("testdata", "portfolio_names_testdata.csv", package = "trisk.analysis"))
```

Leaving the company_id and company_name columns empty, Trisk results will be aggregated per country and technology, and matched to the portfolio based on those columns.

```{r echo=FALSE}
knitr::kable(portfolio_countries_testdata) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

Filling in the company_name column, will result in an attempt to fuzzy string matching between company names.

```{r echo=FALSE}
knitr::kable(portfolio_names_testdata) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

Filling in the company_id column, will result in an exact match between companies.

```{r echo=FALSE}
knitr::kable(portfolio_ids_testdata) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

Using the company ids is recommended to match the portfolio. In our current asset data, a unique asset is defined by a unique combination of company_id, sector, technology, and country. Those other columns are used for the matching between the portfolio and the Trisk outputs.

```{r}
portfolio_testdata <- portfolio_ids_testdata
```


### Run trisk 

Run the model with the provided data, after filtering assets on those available in the portfolio.

Define the scenarios to use:
```{r}
baseline_scenario <- "NGFS2023GCAM_CP"
target_scenario <- "NGFS2023GCAM_NZ2050"
scenario_geography <- "OecdAndEu"
```

The function `run_trisk_on_portfolio()` handles the filtering on portfolio and then runs Trisk:
```{r}
analysis_data <- run_trisk_on_portfolio(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  portfolio_data = portfolio_testdata,
  baseline_scenario = baseline_scenario,
  target_scenario = target_scenario,
  scenario_geography = scenario_geography
)
```

Result dataframe : 

```{r echo=FALSE}
knitr::kable(analysis_data) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

## Plot results

### Equities risk

Plot the average percentage of NPV change per technology

```{r}
pipeline_crispy_npv_change_plot(analysis_data)
```

Plot the resulting portfolio's exposure change 

```{r}
pipeline_crispy_exposure_change_plot(analysis_data)
```
### Bonds&Loans risk

Plot the average PDs at baseline and shock

```{r}
pipeline_crispy_pd_term_plot(analysis_data)
```

Plot the resulting portfolio's expected loss

```{r}
pipeline_crispy_expected_loss_plot(analysis_data)
```
