---
title: "portfolio-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{portfolio-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages({
  suppressWarnings(library(trisk.analysis))
  suppressWarnings(library(magrittr))
})
```

# Restrict the analysis to a portfolio

## Generate outputs
### Load the test data

Load the packaged Khanbank asset and financial inputs together with the local Asia scenario and carbon reference files.

```{r}
project_dir_candidates <- c(getwd(), normalizePath(file.path(getwd(), ".."), mustWork = FALSE))
project_dir <- project_dir_candidates[file.exists(file.path(project_dir_candidates, "DESCRIPTION"))][1]

repo_or_pkg_file <- function(filename) {
  repo_path <- file.path(project_dir, "inst", "testdata", filename)
  if (file.exists(repo_path)) {
    return(repo_path)
  }

  pkg_path <- system.file("testdata", filename, package = "trisk.analysis", mustWork = FALSE)
  if (nzchar(pkg_path) && file.exists(pkg_path)) {
    return(pkg_path)
  }

  stop("Could not locate input file: ", filename)
}

assets_testdata <- read.csv(repo_or_pkg_file("assets_data_khanbank_ifc.csv"))
assets_testdata$company_name <- as.character(assets_testdata$company_name)
assets_testdata$asset_name <- as.character(assets_testdata$asset_name)

scenarios_testdata <- read.csv(repo_or_pkg_file("scenarios_mongolia_client.csv"))
financial_features_testdata <- read.csv(repo_or_pkg_file("financial_features_khanbank_ifc.csv"))[
  c("company_id", "pd", "net_profit_margin", "debt_equity_ratio", "volatility")
]
ngfs_carbon_price_testdata <- read.csv(repo_or_pkg_file("ngfs_carbon_price_mongolia_client.csv"))

output_dir <- file.path(project_dir, "client_outputs", "khanbank-portfolio-analysis")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```


### Prepare portfolio

The Khanbank debug bundle includes two exact-match portfolio input structures:

```{r}
portfolio_detailed_testdata <- read.csv(repo_or_pkg_file("portfolio_khanbank_ifc.csv"))
portfolio_detailed_testdata$company_name <- as.character(portfolio_detailed_testdata$company_name)

portfolio_simple_testdata <- read.csv(repo_or_pkg_file("simple_portfolio_khanbank_ifc.csv"))
portfolio_simple_testdata$company_name <- as.character(portfolio_simple_testdata$company_name)
```

The detailed portfolio keeps the exact `company_id`, `sector`, `technology`, and `country_iso2` fields alongside exposure information.

```{r echo=FALSE}
knitr::kable(portfolio_detailed_testdata) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

The simplified portfolio keeps exact `company_id` matching but omits the sector and technology breakdown.

```{r echo=FALSE}
knitr::kable(portfolio_simple_testdata) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

Using the detailed file is recommended here because the Khanbank asset set is already broken out by company, sector, technology, and country.

```{r}
portfolio_testdata <- portfolio_detailed_testdata
```

The vignette uses the Khanbank portfolio bundle directly so the output reflects the same debug inputs as the assets and financial data.

### Run trisk 

Run the model with the provided data, after filtering assets on those available in the portfolio.

Define a central Khanbank Mongolia case to use:
```{r}
baseline_scenario <- "NGFS2024GCAM_CP"
target_scenario <- "NGFS2024GCAM_DT"
scenario_geography <- "Asia"
risk_free_rate <- 0.04
discount_rate <- 0.10
growth_rate <- 0.03
shock_year <- 2030
carbon_price_model <- "no_carbon_tax"
```

The function `run_trisk_on_portfolio()` handles the filtering on portfolio and then runs Trisk:
```{r}
analysis_data <- run_trisk_on_portfolio(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  portfolio_data = portfolio_testdata,
  baseline_scenario = baseline_scenario,
  target_scenario = target_scenario,
  scenario_geography = scenario_geography,
  risk_free_rate = risk_free_rate,
  discount_rate = discount_rate,
  growth_rate = growth_rate,
  shock_year = shock_year,
  carbon_price_model = carbon_price_model
)

utils::write.csv(
  analysis_data,
  file.path(output_dir, "portfolio_analysis_data.csv"),
  row.names = FALSE
)
```

Result dataframe : 

```{r echo=FALSE}
knitr::kable(analysis_data) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

## Plot results

### Equities risk

Plot the average percentage of NPV change per technology

```{r fig.width=10, fig.height=6}
npv_change_plot <- pipeline_crispy_npv_change_plot(analysis_data)
npv_change_plot
ggplot2::ggsave(
  filename = file.path(output_dir, "npv_change_plot.png"),
  plot = npv_change_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

Plot the resulting portfolio's exposure change 

```{r fig.width=10, fig.height=6}
exposure_change_plot <- pipeline_crispy_exposure_change_plot(analysis_data)
exposure_change_plot
ggplot2::ggsave(
  filename = file.path(output_dir, "exposure_change_plot.png"),
  plot = exposure_change_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```
### Bonds&Loans risk

Plot the average PDs at baseline and shock

```{r fig.width=10, fig.height=7}
pd_term_plot <- pipeline_crispy_pd_term_plot(analysis_data)
pd_term_plot
ggplot2::ggsave(
  filename = file.path(output_dir, "pd_term_plot.png"),
  plot = pd_term_plot,
  width = 10,
  height = 7,
  dpi = 300
)
```

Plot the resulting portfolio's expected loss

```{r fig.width=10, fig.height=6}
expected_loss_plot <- pipeline_crispy_expected_loss_plot(analysis_data)
expected_loss_plot
ggplot2::ggsave(
  filename = file.path(output_dir, "expected_loss_plot.png"),
  plot = expected_loss_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

Saved files:

```{r}
list.files(output_dir)
```
