---
title: "3. Run on a portfolio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Run on a portfolio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(dplyr)
library(magrittr)
```

## Overview

This vignette walks through running TRISK on a bank loan book. The package
exposes two entry points, and the right one depends on how much company
metadata your portfolio carries:

- `run_trisk_on_simple_portfolio()` — the **minimal path**. You bring company
  exposures, terms, and loss-given-default; no `country_iso2` column is needed.
  TRISK runs and allocates exposure across each company's technologies for you.
- `run_trisk_on_portfolio()` — the **fuller path**. You match your portfolio to
  TRISK assets by country and company (by ID, name, or country aggregate), then
  run the model on the filtered asset universe.

As a rule of thumb: use the simple runner for a quick, single-company-resolution
risk read where you already know which counterparties you hold. Use the full
runner when you need country-level matching, fuzzy name matching, or the
built-in equity- and credit-risk plots.

## Inputs

Both runners need the same four TRISK model inputs, shipped as test data in the
`trisk.model` package:

```{r}
assets_testdata <- read.csv(system.file("testdata", "assets_testdata.csv", package = "trisk.model", mustWork = TRUE))
scenarios_testdata <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model", mustWork = TRUE))
financial_features_testdata <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model", mustWork = TRUE))
ngfs_carbon_price_testdata <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model", mustWork = TRUE))
```

They also share the same scenario settings — a baseline and a target scenario,
plus the geography to evaluate:

```{r}
baseline_scenario <- "NGFS2023GCAM_CP"
target_scenario <- "NGFS2023GCAM_NZ2050"
scenario_geography <- "Global"
```

The two runners differ only in the **portfolio schema** they expect, covered in
each example below.

## Minimal example

`run_trisk_on_simple_portfolio()` expects a portfolio with just five columns:

- `company_id`
- `company_name`
- `exposure_value_usd`
- `term`
- `loss_given_default`

No `country_iso2` is required. Load the bundled simple portfolio:

```{r}
simple_portfolio <- read.csv(
  system.file("testdata", "simple_portfolio.csv", package = "trisk.analysis", mustWork = TRUE)
)
simple_portfolio
```

Run the model:

```{r}
simple_results <- run_trisk_on_simple_portfolio(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  portfolio_data = simple_portfolio,
  baseline_scenario = baseline_scenario,
  target_scenario = target_scenario,
  scenario_geography = scenario_geography
)

portfolio_results_tech_detail <- simple_results$portfolio_results_tech_detail
portfolio_results <- simple_results$portfolio_results
```

### NPV-based exposure allocation

The simple runner adds a column the full runner does not: `exposure_value_usd_share`.
A single loan exposure has to be spread across the company's technologies before
technology-level risk can be attributed to it. TRISK allocates exposure from
baseline NPV shares at company/sector/technology level:

1. compute baseline NPV share per run;
2. allocate exposure with that share;
3. average after dropping `run_id`;
4. re-scale so exposure totals match the original portfolio exposure.

```{r}
portfolio_results_tech_detail |>
  dplyr::select(
    company_id, term, sector, technology,
    exposure_value_usd_share,
    net_present_value_baseline
  ) |>
  utils::head(10)
```

Because `exposure_value_usd_share` is computed after dropping `run_id`, it is
constant across runs for a given `(company_id, term, sector, technology)`.

```{r}
exposure_share_check <- portfolio_results_tech_detail |>
  dplyr::distinct(
    company_id, term, sector, technology,
    exposure_value_usd_share
  ) |>
  dplyr::group_by(company_id, term) |>
  dplyr::summarise(allocated_exposure = sum(exposure_value_usd_share, na.rm = TRUE), .groups = "drop") |>
  dplyr::left_join(
    portfolio_results |>
      dplyr::group_by(company_id, term) |>
      dplyr::summarise(original_exposure = sum(exposure_value_usd, na.rm = TRUE), .groups = "drop"),
    by = c("company_id", "term")
  ) |>
  dplyr::mutate(gap = allocated_exposure - original_exposure)

exposure_share_check
```

The `gap` column should be close to zero (floating-point tolerance) — confirming
allocated exposure reconciles back to the original loan amounts.

## Full portfolio with country matching

When your portfolio carries country information, `run_trisk_on_portfolio()`
filters TRISK assets to your holdings before running, and gives you matching
flexibility. There are three valid portfolio input structures:

```{r}
portfolio_countries_testdata <- read.csv(system.file("testdata", "portfolio_countries_testdata.csv", package = "trisk.analysis"))
portfolio_ids_testdata <- read.csv(system.file("testdata", "portfolio_ids_testdata.csv", package = "trisk.analysis"))
portfolio_names_testdata <- read.csv(system.file("testdata", "portfolio_names_testdata.csv", package = "trisk.analysis"))
```

Leaving `company_id` and `company_name` empty aggregates TRISK results per
country and technology, matched on those columns:

```{r echo=FALSE}
knitr::kable(portfolio_countries_testdata) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

Filling in `company_name` triggers fuzzy string matching between company names:

```{r echo=FALSE}
knitr::kable(portfolio_names_testdata) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

Filling in `company_id` triggers an exact match between companies:

```{r echo=FALSE}
knitr::kable(portfolio_ids_testdata) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

Using company IDs is the recommended match. In the asset data, a unique asset is
defined by a unique combination of `company_id`, `sector`, `technology`, and
`country`. Those columns drive the join between the portfolio and TRISK outputs:

```{r}
portfolio_testdata <- portfolio_ids_testdata
```

`run_trisk_on_portfolio()` handles the filtering on portfolio and then runs TRISK:

```{r}
analysis_data <- run_trisk_on_portfolio(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  portfolio_data = portfolio_testdata,
  baseline_scenario = baseline_scenario,
  target_scenario = target_scenario,
  scenario_geography = scenario_geography
)
```

Result data frame:

```{r echo=FALSE}
knitr::kable(analysis_data) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

### Plotting the results

The full runner's output plugs directly into the package plotting helpers.

**Equity risk** — average percentage NPV change per technology:

```{r}
pipeline_crispy_npv_change_plot(analysis_data)
```

Resulting portfolio exposure change:

```{r}
pipeline_crispy_exposure_change_plot(analysis_data)
```

**Bonds & loans risk** — average PDs at baseline and shock:

```{r}
pipeline_crispy_pd_term_plot(analysis_data)
```

Resulting portfolio expected loss:

```{r}
pipeline_crispy_expected_loss_plot(analysis_data)
```

## Interpretation

- The **simple runner** is the fastest way to attribute climate transition risk
  to a known book. Its `exposure_value_usd_share` lets you see how each loan's
  exposure spreads across a company's technologies, so you can spot which
  technology mix drives a counterparty's risk.
- The **full runner** is the right tool when matching is non-trivial — you only
  have country-level exposures, or you need to reconcile messy company names —
  and when you want the equity (NPV change, exposure change) and credit (PD term,
  expected loss) views straight out of the box.
- A negative NPV change and a rising shock PD on the same technology flag where
  the transition both erodes asset value and lifts default probability — the
  positions to scrutinise first.

## Caveats

- The country-aggregate and fuzzy-name matching modes trade precision for
  coverage. Prefer `company_id` exact matching whenever IDs are available; fall
  back to name or country matching only when they are not.
- Exposure allocation in the simple runner is NPV-share based and reconciles to
  the original totals only up to floating-point tolerance — confirm the `gap`
  before relying on allocated figures.
- The scenarios, geography, and bundled test data here are illustrative. Swap in
  your own asset, scenario, financial, and carbon-price inputs for production
  analysis.

## See also

- `getting-started` — install and run a first TRISK analysis.
- `inputs-and-outputs` — the input schemas and output tables in detail.
- `pd-el-integration` — turning TRISK output into PD and expected-loss measures.
- `sensitivity-analysis` — varying scenarios and parameters across runs.
