---
title: "sensitivity-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sensitivity-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages({
  suppressWarnings(library(trisk.analysis))
  suppressWarnings(library(magrittr))
  suppressWarnings(library(dplyr))
  suppressWarnings(library(tidyr))
  suppressWarnings(library(ggplot2))
  suppressWarnings(library(scales))
})
```

```{r}
project_dir_candidates <- c(getwd(), normalizePath(file.path(getwd(), ".."), mustWork = FALSE))
project_dir <- project_dir_candidates[file.exists(file.path(project_dir_candidates, "DESCRIPTION"))][1]

repo_or_pkg_file <- function(filename) {
  repo_path <- file.path(project_dir, "inst", "testdata", filename)
  if (file.exists(repo_path)) {
    return(repo_path)
  }

  pkg_path <- system.file("testdata", filename, package = "trisk.analysis", mustWork = FALSE)
  if (nzchar(pkg_path) && file.exists(pkg_path)) {
    return(pkg_path)
  }

  stop("Could not locate input file: ", filename)
}

assets_testdata <- read.csv(repo_or_pkg_file("assets_data_khanbank_ifc.csv"))
assets_testdata$company_name <- as.character(assets_testdata$company_name)
assets_testdata$asset_name <- as.character(assets_testdata$asset_name)

scenarios_testdata <- read.csv(repo_or_pkg_file("scenarios_mongolia_client.csv"))
financial_features_testdata <- read.csv(repo_or_pkg_file("financial_features_khanbank_ifc.csv"))[
  c("company_id", "pd", "net_profit_margin", "debt_equity_ratio", "volatility")
]
ngfs_carbon_price_testdata <- read.csv(repo_or_pkg_file("ngfs_carbon_price_mongolia_client.csv"))

output_dir <- file.path(project_dir, "client_outputs", "khanbank-sensitivity-analysis")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

# Sensitivity Analysis

This vignette runs a Khanbank-focused sensitivity grid on the collapsed Mongolia client asset dataset.

## Run matrix

The requested setup is:

- Asia geography only
- NGFS 2024 `GCAM`, `MESSAGE`, and `REMIND`
- baseline `CP` and target `DT`
- shock years `2026`, `2030`, and `2035`
- discount rates `10%` and `15%`
- growth rate `3%`
- risk-free rates `4%` and `5%`

Because `shock_year` is scalar per TRISK run, the three shock years are represented as separate entries in the sensitivity grid rather than passed as one vector to a single run.

```{r}
run_grid <- tidyr::expand_grid(
  model_family = c("GCAM", "MESSAGE", "REMIND"),
  shock_year = c(2026, 2030, 2035),
  discount_rate = c(0.10, 0.15),
  risk_free_rate = c(0.04, 0.05)
) %>%
  mutate(
    scenario_geography = "Asia",
    baseline_scenario = paste0("NGFS2024", .data$model_family, "_CP"),
    target_scenario = paste0("NGFS2024", .data$model_family, "_DT"),
    growth_rate = 0.03,
    carbon_price_model = "no_carbon_tax",
    div_netprofit_prop_coef = 1,
    market_passthrough = 0
  ) %>%
  relocate(
    model_family,
    baseline_scenario,
    target_scenario,
    scenario_geography
  )

run_params <- run_grid %>%
  select(
    -model_family
  ) %>%
  split(seq_len(nrow(.))) %>%
  lapply(as.list)

knitr::kable(run_grid) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

## Run sensitivity analysis

```{r}
sensitivity_analysis_results <- run_trisk_sa(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  run_params = run_params
)
```

## Summarise results by run

```{r}
npv_summary <- sensitivity_analysis_results$npv %>%
  group_by(.data$run_id) %>%
  summarise(
    total_npv_baseline = sum(.data$net_present_value_baseline, na.rm = TRUE),
    total_npv_shock = sum(.data$net_present_value_shock, na.rm = TRUE),
    total_npv_difference = sum(.data$net_present_value_difference, na.rm = TRUE),
    total_npv_change = .data$total_npv_difference / .data$total_npv_baseline,
    .groups = "drop"
  )

pd_summary <- sensitivity_analysis_results$pd %>%
  filter(.data$term == 5) %>%
  group_by(.data$run_id) %>%
  summarise(
    median_pd_baseline_term5 = median(.data$pd_baseline, na.rm = TRUE),
    median_pd_shock_term5 = median(.data$pd_shock, na.rm = TRUE),
    median_pd_difference_term5 = median(.data$pd_shock - .data$pd_baseline, na.rm = TRUE),
    .groups = "drop"
  )

run_summary <- sensitivity_analysis_results$params %>%
  left_join(npv_summary, by = "run_id") %>%
  left_join(pd_summary, by = "run_id") %>%
  mutate(
    model_family = sub("^NGFS2024([A-Z]+)_CP$", "\\1", .data$baseline_scenario),
    discount_rate_label = paste0("Discount ", percent(.data$discount_rate, accuracy = 1)),
    risk_free_rate_label = paste0("Risk-free ", percent(.data$risk_free_rate, accuracy = 1)),
    run_label = paste(.data$model_family, "shock", .data$shock_year)
  ) %>%
  arrange(.data$model_family, .data$discount_rate, .data$risk_free_rate, .data$shock_year)

utils::write.csv(
  run_summary,
  file.path(output_dir, "run_summary.csv"),
  row.names = FALSE
)
utils::write.csv(
  sensitivity_analysis_results$params,
  file.path(output_dir, "run_params_used.csv"),
  row.names = FALSE
)

knitr::kable(run_summary) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

## Visualisations

### Total NPV change by model family and shock year

```{r fig.width=12, fig.height=7}
npv_sensitivity_plot <- ggplot(
  run_summary,
  aes(
    x = factor(.data$shock_year),
    y = .data$total_npv_change,
    color = .data$model_family,
    group = .data$model_family
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.4) +
  facet_wrap(~discount_rate_label) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Khanbank Sensitivity: Total NPV Change",
    subtitle = "Asia geography, NGFS 2024 CP baseline vs DT shock",
    x = "Shock year",
    y = "Total portfolio NPV change",
    color = "Model family"
  )

npv_sensitivity_plot

ggplot2::ggsave(
  filename = file.path(output_dir, "npv_sensitivity_plot.png"),
  plot = npv_sensitivity_plot,
  width = 12,
  height = 7,
  dpi = 300
)
```

### Term-5 median PD shock by model family and shock year

```{r fig.width=13, fig.height=8}
pd_sensitivity_plot <- ggplot(
  run_summary,
  aes(
    x = factor(.data$shock_year),
    y = .data$median_pd_shock_term5,
    color = .data$model_family,
    group = .data$model_family
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.4) +
  facet_grid(
    rows = vars(risk_free_rate_label),
    cols = vars(discount_rate_label)
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Khanbank Sensitivity: Term-5 Median PD Shock",
    subtitle = "Asia geography, NGFS 2024 CP baseline vs DT shock",
    x = "Shock year",
    y = "Median term-5 PD under shock",
    color = "Model family"
  )

pd_sensitivity_plot

ggplot2::ggsave(
  filename = file.path(output_dir, "pd_sensitivity_plot.png"),
  plot = pd_sensitivity_plot,
  width = 13,
  height = 8,
  dpi = 300
)
```

### Central-assumption production trajectory comparison

This plot focuses on the lower-end financing assumption (`discount_rate = 10%`, `risk_free_rate = 4%`) to keep the trajectory comparison readable.

```{r fig.width=12, fig.height=10}
central_runs <- run_summary %>%
  filter(.data$discount_rate == 0.10, .data$risk_free_rate == 0.04) %>%
  select(.data$run_id, .data$model_family, .data$shock_year, .data$run_label)

central_trajectories <- sensitivity_analysis_results$trajectories %>%
  inner_join(central_runs, by = "run_id")

central_trajectory_plot <- plot_multi_trajectories(
  trajectories_data = central_trajectories,
  facet_var = "technology",
  linecolor = "run_label"
)

central_trajectory_plot

ggplot2::ggsave(
  filename = file.path(output_dir, "central_trajectory_plot.png"),
  plot = central_trajectory_plot,
  width = 12,
  height = 10,
  dpi = 300
)
```

## Saved files

```{r}
list.files(output_dir)
```
