---
title: "3. Inputs and outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Inputs and outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
is_CRAN <- !identical(Sys.getenv("NOT_CRAN"), "true")
```

```{r setup}
library(trisk.analysis)
library(dplyr)
library(magrittr)
```

## Overview

This vignette is the reference for *what TRISK consumes and what it produces*.
TRISK is a climate transition stress test: it compares each company's value and
default risk under a baseline scenario against a more ambitious (shock) scenario,
so a credit analyst can see how a low-carbon transition reprices a book of
counterparties. To run it you supply four reference datasets plus the model
parameters; you get back three result tables (NPV, PD, and company
trajectories). The sections below enumerate each input schema, show how to obtain
production inputs, run a minimal example on the bundled test data, and explain
how to read every output column.

## Inputs

> **Input data — where your data goes.** TRISK needs **five inputs**: four that
> describe the world — **assets**, **scenarios**, **NGFS carbon price** and
> **financial features** — plus your **portfolio**. The main portfolio file is
> **`portfolio_ids`** (matched by `company_id`); `portfolio_names` and
> `portfolio_countries` are options. **The CSVs loaded below are placeholders**
> (bundled samples) — replace them with your own files. See
> [Bank portfolio analysis](v2_bank-portfolio-analysis.html) for `setup_trisk_inputs()`
> and the `trisk_inputs/` folder convention.

TRISK requires four reference datasets, all shipped as test data inside
`trisk.model`. In production you swap these for the downloaded inputs (see
*Obtaining inputs* below) and for your own portfolio file (a list of
`company_id`s you hold exposure to, used downstream to filter results).

| Input | Argument | Grain | Purpose |
|---|---|---|---|
| Assets | `assets_data` | company / asset / year | Physical production capacity and emissions |
| Scenarios | `scenarios_data` | scenario / sector / technology / year | Price and production pathways per climate scenario |
| Financial features | `financial_data` | company | Starting PD and balance-sheet ratios |
| NGFS carbon price | `carbon_data` | model / scenario / year | Optional carbon tax trajectories |
| Portfolio | (downstream) | company | The counterparties you hold; filters results |

Load the bundled test datasets:

```{r}
assets_testdata <- read.csv(system.file("testdata", "assets_testdata.csv", package = "trisk.model"))
scenarios_testdata <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model"))
financial_features_testdata <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model"))
ngfs_carbon_price_testdata <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model"))
```

### Assets

Company production assets: capacity, technology, sector and location. This is the
physical exposure TRISK reprices.

- `company_id`, `company_name`: company identifiers.
- `asset_id`, `asset_name`: asset identifiers.
- `country_iso2`: ISO 3166-1 alpha-2 country code.
- `production_year`: year of the production record.
- `emission_factor`: emissions per unit of production.
- `technology`, `sector`: technology and production sector.
- `capacity`, `capacity_factor`: installed capacity and utilisation.
- `production_unit`: unit of production.

```{r}
str(assets_testdata)
```

```{r echo=FALSE}
knitr::kable(head(assets_testdata)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px") %>%
  kableExtra::column_spec(1:ncol(assets_testdata), width = "150px")
```

### Scenarios

Price and production pathways per climate scenario. Each TRISK run picks one
baseline and one target scenario from this table.

- `scenario_geography`: region the pathway applies to.
- `scenario`, `scenario_pathway`, `scenario_type`: scenario name, pathway and type (baseline or shock).
- `sector`, `technology`, `technology_type`: technology classification (carbon or renewable).
- `scenario_year`: year of the pathway record.
- `scenario_price`, `price_unit`, `pathway_unit`: price and units.
- `scenario_capacity_factor`, `country_iso2_list`, `scenario_provider`: capacity factor, the countries the geography aggregates, and the IAM/source that produced the pathway.

```{r}
str(scenarios_testdata)
```

```{r echo=FALSE}
knitr::kable(head(scenarios_testdata, 50)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "200%", height = "400px")
```

### Financial features

Per-company starting financials. The `pd` here is the pre-stress probability of
default that TRISK shocks.

- `company_id`: company identifier.
- `pd`: starting probability of default.
- `net_profit_margin`, `debt_equity_ratio`, `volatility`: balance-sheet inputs to the valuation and Merton PD model.

```{r}
str(financial_features_testdata)
```

```{r echo=FALSE}
knitr::kable(head(financial_features_testdata)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

### NGFS carbon price

Optional carbon tax trajectories applied to company profits. Used only when you
select a `carbon_price_model` other than `no_carbon_tax`.

- `year`, `model`, `scenario`: carbon price record keys.
- `scenario_geography`, `variable`, `unit`: region and units.
- `carbon_tax`: tax amount per ton of CO2.

```{r}
str(ngfs_carbon_price_testdata)
```

```{r echo=FALSE}
knitr::kable(head(ngfs_carbon_price_testdata)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

### Obtaining inputs

The test data above is small and fixed. For real analysis, download the full
production inputs (assets, scenarios, financial features and carbon prices) from
the public endpoint:

```{r}
print(
  paste0(trisk.analysis:::TRISK_DATA_INPUT_ENDPOINT, "/", trisk.analysis:::TRISK_DATA_S3_PREFIX)
)
```

`download_trisk_inputs()` fetches the production `scenarios.csv` into a folder of
your choice (skipped here so the vignette has no network dependency). The other
three reference inputs — `assets`, `financial_features` and `ngfs_carbon_price` —
you supply yourself, starting from the bundled samples (see
[Bank portfolio analysis](v2_bank-portfolio-analysis.html) and `setup_trisk_inputs()`):

```{r}
trisk_inputs_folder <- file.path(".", "trisk_inputs")
if (!is_CRAN) {
  download_trisk_inputs(local_save_folder = trisk_inputs_folder, skip_confirmation = TRUE)
}
```

Once downloaded you read the CSVs the same way as the test data. The production
`scenarios.csv` covers many more scenarios than the bundled sample: six
providers (NGFS, WEO, IPR, Oxford, GECO and the Mission Possible Partnership for
steel), each with one or more data vintages.

Scenario names encode provider, vintage and scenario, e.g. `WEO2023_STEPS` is the
IEA World Energy Outlook 2023 "Stated Policies" baseline; `NGFS2023GCAM_NZ2050`
is the NGFS 2023 "Net Zero 2050" target run on the GCAM integrated-assessment
model. As a rule, pick the baseline and target from the *same provider*.

The bundled scenarios already expose the valid baseline/target pairings via
`get_available_parameters()` (run here on the test data so it needs no network):

```{r}
available_parameters <- get_available_parameters(scenarios_testdata)
```

```{r echo=FALSE}
knitr::kable(available_parameters) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

When choosing a scenario consider: data vintage (newer is usually more relevant),
geography coverage (NGFS GCAM offers sub-global regions), the target's ambition
level (e.g. NGFS B2DS below 2 degrees vs NZ2050 net zero), and narrative — some
NGFS scenarios such as "LD" (Low Demand) reach net zero through demand assumptions
rather than policy alone. See <https://www.ngfs.net/en> for narrative detail.

## Minimal example

The smallest runnable stress test against the bundled data. The parameters fall
into two groups: scenario selection (which pathways to compare and where) and
financial assumptions (the DCF and PD model knobs).

```{r}
baseline_scenario <- "NGFS2023GCAM_CP"
target_scenario <- "NGFS2023GCAM_NZ2050"
scenario_geography <- "Global"
shock_year <- 2030

carbon_price_model <- "no_carbon_tax"
risk_free_rate <- 0.02
discount_rate <- 0.07
growth_rate <- 0.03
div_netprofit_prop_coef <- 1
market_passthrough <- 0
```

Parameter reference:

- `baseline_scenario`, `target_scenario`: the status-quo and shock scenario names.
- `scenario_geography`: region(s) to compute. Choosing a region other than
  `"Global"` filters assets to that region's countries and uses region-specific
  pathways.
- `shock_year`: the year the transition shock is applied.
- `carbon_price_model`: which carbon tax trajectory to apply (`no_carbon_tax` to skip).
- `market_passthrough`: share of the carbon tax the firm passes to customers.
- `risk_free_rate`, `discount_rate`, `growth_rate`, `div_netprofit_prop_coef`:
  DCF and PD valuation assumptions.

`run_trisk_model()` runs the model and returns results at the asset grain
(one row per company / asset / country / technology):

```{r}
st_results <- run_trisk_model(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  baseline_scenario = baseline_scenario,
  target_scenario = target_scenario,
  scenario_geography = scenario_geography,
  shock_year = shock_year,
  carbon_price_model = carbon_price_model,
  risk_free_rate = risk_free_rate,
  discount_rate = discount_rate,
  growth_rate = growth_rate,
  div_netprofit_prop_coef = div_netprofit_prop_coef,
  market_passthrough = market_passthrough
)

npv_results <- st_results$npv_results
pd_results <- st_results$pd_results
company_trajectories <- st_results$company_trajectories
```

### Company-aggregated results

Since December 2024 TRISK runs at the company / country / technology grain, which
improves NPV and trajectory accuracy but changes the output shape relative to
older analyses. `run_trisk_agg()` takes the same arguments and aggregates NPVs
per company over country and technology, restoring the pre-2024 output structure
for backward compatibility or simpler reporting. PD results are unaffected — they
are already at the company/sector grain in both functions.

```{r}
st_results_agg <- run_trisk_agg(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  baseline_scenario = baseline_scenario,
  target_scenario = target_scenario,
  scenario_geography = scenario_geography,
  shock_year = shock_year
)

npv_results_agg <- st_results_agg$npv_results
```

The aggregated NPV table drops the `country_iso2` column:

```{r echo=FALSE}
knitr::kable(head(as.data.frame(npv_results_agg))) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

### Adding a carbon tax

To stress profits with a carbon tax, pick a `carbon_price_model` present in your
carbon data and a `market_passthrough` between 0 and 1. The test data ships the
`GCAM 5.3+ NGFS` trajectory plus several illustrative flat and increasing taxes:

```{r}
ngfs_carbon_price_testdata %>%
  distinct(model)
```

```{r}
st_results_tax <- run_trisk_model(
  assets_data = assets_testdata,
  scenarios_data = scenarios_testdata,
  financial_data = financial_features_testdata,
  carbon_data = ngfs_carbon_price_testdata,
  baseline_scenario = baseline_scenario,
  target_scenario = target_scenario,
  scenario_geography = scenario_geography,
  carbon_price_model = "GCAM 5.3+ NGFS",
  market_passthrough = 0.3
)
```

## Outputs and interpretation

Each run returns a list of three tables. For a credit analyst the headline is the
*difference* between baseline and shock columns: a large drop from baseline to
shock NPV, or a jump from baseline to shock PD, flags a counterparty whose value
or creditworthiness is sensitive to the transition.

### NPV results

Net present value of each asset under both scenarios. Compare
`net_present_value_baseline` against `net_present_value_shock`: the gap is the
transition value-at-risk for that asset.

- `run_id`: identifier for the simulation run.
- `company_id`, `asset_id`, `company_name`, `asset_name`: identifiers.
- `country_iso2`: country of the asset (absent in `run_trisk_agg()` output).
- `sector`, `technology`: classification (e.g. Oil&Gas, Coal, Power; Gas, CoalCap, RenewablesCap).
- `net_present_value_baseline`, `net_present_value_shock`: NPV under each scenario.
- `net_present_value_difference`, `net_present_value_change`: shock minus baseline (absolute) and the relative change.

```{r}
str(npv_results)
```

```{r echo=FALSE}
knitr::kable(head(as.data.frame(npv_results))) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

### PD results

Probability of default per company and term, before and after the shock. The
`pd_shock - pd_baseline` spread is the transition-driven increase in default risk
to feed into credit models.

- `run_id`, `company_id`, `company_name`: identifiers.
- `sector`: company sector.
- `term`: horizon of the PD estimate.
- `pd_baseline`, `pd_shock`: PD under each scenario.

```{r}
str(pd_results)
```

```{r echo=FALSE}
knitr::kable(head(as.data.frame(pd_results))) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

### Company trajectories

The year-by-year detail behind the NPV and PD numbers: production, prices and
profits along each pathway. Use this to audit *why* a company's value moved — for
example a coal asset whose shock-scenario production collapses while its baseline
holds.

- `run_id`, `asset_id`, `asset_name`, `company_id`, `company_name`, `country_iso2`: identifiers.
- `year`, `sector`, `technology`: time and classification.
- `production_plan_company_technology`: the company's own production plan.
- `production_baseline_scenario`, `production_target_scenario`, `production_shock_scenario`: production along each pathway.
- `pd`, `net_profit_margin`, `debt_equity_ratio`, `volatility`: carried-through financial features.
- `scenario_price_baseline`, `price_shock_scenario`: prices per scenario.
- `net_profits_baseline_scenario`, `net_profits_shock_scenario`: net profits per scenario.
- `discounted_net_profits_baseline_scenario`, `discounted_net_profits_shock_scenario`: discounted net profits feeding the NPV.

```{r}
str(company_trajectories)
```

```{r echo=FALSE}
knitr::kable(head(as.data.frame(company_trajectories))) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

## Caveats

- **Scenario coverage is uneven.** Some sectors (notably steel) have far fewer
  scenarios than power, because of upstream data limitations. Always pair a
  baseline and target from the same provider and vintage.
- **Geography filtering is implicit.** Choosing a `scenario_geography` other than
  `"Global"` silently drops assets outside that region's countries — check your
  asset count after filtering.
- **Carbon tax requires matching data.** A `carbon_price_model` only takes effect
  if that model exists in `carbon_data`; otherwise leave it as `no_carbon_tax`.
- **Test data is illustrative.** The bundled datasets exist to make examples
  runnable, not to produce representative risk numbers — use the downloaded
  production inputs for real analysis.

## See also

- `getting-started` — install and run your first stress test.
- `run-on-a-portfolio` — apply TRISK to a portfolio of counterparties.
- `pd-el-integration` — turn shocked PDs into expected-loss figures.
- `sensitivity-analysis` — sweep parameters to test result stability.
