---
title: "pd-el-integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pd-el-integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trisk.analysis)
library(magrittr)
```

# PD and EL Integration

## 1. Setup

Load the bundled testdata shipped with `trisk.model` and `trisk.analysis`.

```{r}
assets_testdata    <- read.csv(system.file("testdata", "assets_testdata.csv",    package = "trisk.model"))
scenarios_testdata <- read.csv(system.file("testdata", "scenarios_testdata.csv", package = "trisk.model"))
fin_testdata       <- read.csv(system.file("testdata", "financial_features_testdata.csv", package = "trisk.model"))
carbon_testdata    <- read.csv(system.file("testdata", "ngfs_carbon_price_testdata.csv", package = "trisk.model"))
portfolio_testdata <- read.csv(system.file("testdata", "portfolio_ids_testdata.csv", package = "trisk.analysis"))
```

## 2. Run TRISK on the portfolio

```{r}
analysis_data <- run_trisk_on_portfolio(
  assets_data       = assets_testdata,
  scenarios_data    = scenarios_testdata,
  financial_data    = fin_testdata,
  carbon_data       = carbon_testdata,
  portfolio_data    = portfolio_testdata,
  baseline_scenario = "NGFS2023GCAM_CP",
  target_scenario   = "NGFS2023GCAM_NZ2050"
)
```

```{r echo=FALSE}
knitr::kable(head(analysis_data[, c("company_id", "sector", "technology", "term",
                                    "pd_baseline", "pd_shock",
                                    "net_present_value_baseline", "net_present_value_shock")])) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "300px")
```

## 3. Why integration matters

TRISK recomputes PD from a Merton structural credit model. That PD level is not
directly comparable to the internal PD your institution already uses; only the
change from baseline to shock carries meaning. Three integration methods
translate the TRISK shift into your internal PD scale.

## 4. Method 1 - Absolute

```{r}
result_abs <- integrate_pd(analysis_data, method = "absolute")
```

```{r echo=FALSE}
knitr::kable(head(result_abs$portfolio[, c("company_id", "sector", "internal_pd",
                                           "pd_baseline", "pd_shock",
                                           "trisk_adjusted_pd", "pd_adjustment")])) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "300px")
```

## 5. Method 2 - Relative

> **Note:** when `pd_baseline = 0` the relative method returns the internal PD
> unchanged; the shock signal is lost on zero-baseline rows. Use `absolute` or
> `zscore` if this matters.

```{r}
result_rel <- integrate_pd(analysis_data, method = "relative")
```

## 6. Method 3 - Z-score (Basel IRB)

The z-score method combines PDs in the normal-quantile space, preserving the
non-linear relationship at distribution tails. Recommended for Basel IRB-aligned
institutions.

```{r}
result_zs <- integrate_pd(analysis_data, method = "zscore")
```

## 7. Supplying your own internal PDs

Pass a vector (length = `nrow(analysis_data)`) or a data frame with
`company_id` and `internal_pd`. Unmatched rows fall back to `pd_baseline`.

```{r}
flat_internal <- rep(0.03, nrow(analysis_data))
result_custom <- integrate_pd(analysis_data,
                              internal_pd = flat_internal,
                              method      = "zscore")
```

## 8. Method comparison

```{r fig.width=7, fig.height=4}
pipeline_crispy_pd_method_comparison(analysis_data)
```

## 9. Integration charts

```{r fig.width=7, fig.height=4}
pipeline_crispy_pd_integration_bars(result_zs)
```

## 10. EL integration

`compute_analysis_metrics` adds `expected_loss_baseline` and `expected_loss_shock`
(EAD * PD) which `integrate_el` requires.

```{r}
analysis_data_el <- trisk.analysis:::compute_analysis_metrics(analysis_data)
result_el <- integrate_el(analysis_data_el, method = "relative")
```

```{r fig.width=7, fig.height=4}
pipeline_crispy_el_adjustment_bars(result_el)
```

## 11. Portfolio-level KPIs

```{r}
pipeline_crispy_pd_kpi_table(result_zs$aggregate)
```

```{r}
pipeline_crispy_el_kpi_table(result_el$aggregate)
```

## 12. Sector breakdown

```{r}
pipeline_crispy_el_sector_breakdown_table(result_el$portfolio)
```
