---
title: "Khan Bank Sensitivity Analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
    self_contained: true
date: "2026-04-22"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
```

```{r setup}
devtools::load_all(".")
library(ggplot2); library(dplyr); library(readr)
```

Mirrors `vignettes/sensitivity-analysis.Rmd` applied to Khan Bank's assets
and financials. Varies `shock_year` between 2025 and 2030; all other
parameters held fixed. Scenarios and carbon price are bundled NGFS data
(non-client).

# 1. Load inputs

```{r}
INPUT_DIR   <- "/Users/jakub/Downloads/04.17 debug khanbank"
OUTPUT_DIR  <- "."
TM_TESTDATA <- system.file("testdata", package = "trisk.model")

assets_data    <- read_csv(file.path(INPUT_DIR, "assets_data_final_ifc.csv"),
                           show_col_types = FALSE)
financial_data <- read_csv(file.path(INPUT_DIR, "financial_features_final_ifc.csv"),
                           show_col_types = FALSE) |>
  select(company_id, pd, net_profit_margin, debt_equity_ratio, volatility) |>
  filter(!is.na(company_id))
scenarios_data <- read_csv(file.path(TM_TESTDATA, "scenarios_testdata.csv"),
                           show_col_types = FALSE)
carbon_data    <- read_csv(file.path(TM_TESTDATA, "ngfs_carbon_price_testdata.csv"),
                           show_col_types = FALSE)

cat(
  "Khan Bank assets:       ", nrow(assets_data), "rows,",
    length(unique(assets_data$company_id)), "companies\n",
  "Khan Bank financials:   ", nrow(financial_data), "rows\n",
  "Bundled scenarios:      ", length(unique(scenarios_data$scenario)), "unique\n",
  "Bundled carbon prices:  ", nrow(carbon_data), "rows\n",
  sep = ""
)
```

# 2. Run parameters

Two TRISK runs, differing only in `shock_year`.

```{r}
run_params <- list(
  list(
    scenario_geography = "Global",
    baseline_scenario  = "NGFS2023GCAM_CP",
    target_scenario    = "NGFS2023GCAM_NZ2050",
    shock_year         = 2030
  ),
  list(
    scenario_geography = "Global",
    baseline_scenario  = "NGFS2023GCAM_CP",
    target_scenario    = "NGFS2023GCAM_NZ2050",
    shock_year         = 2025
  )
)
```

# 3. Run the sensitivity analysis

```{r}
sa <- run_trisk_sa(
  assets_data    = assets_data,
  scenarios_data = scenarios_data,
  financial_data = financial_data,
  carbon_data    = carbon_data,
  run_params     = run_params
)
```

# 4. Consolidated per-run outputs

## 4a. NPV

```{r}
knitr::kable(head(sa$npv, 20), caption = "NPV per run (first 20 rows)") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  kableExtra::scroll_box(width = "100%", height = "400px")
write_csv(sa$npv, file.path(OUTPUT_DIR, "sensitivity_npv.csv"))
```

## 4b. PD

```{r}
knitr::kable(head(sa$pd, 20), caption = "PD per run (first 20 rows)") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  kableExtra::scroll_box(width = "100%", height = "400px")
write_csv(sa$pd, file.path(OUTPUT_DIR, "sensitivity_pd.csv"))
```

## 4c. Parameters

```{r}
knitr::kable(sa$params, caption = "Run parameters") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
write_csv(sa$params, file.path(OUTPUT_DIR, "sensitivity_params.csv"))
```

```{r, include = FALSE}
write_csv(sa$trajectories, file.path(OUTPUT_DIR, "sensitivity_trajectories.csv"))
```

# 5. Multi-trajectory plot

```{r fig.width = 10, fig.height = 6}
p <- plot_multi_trajectories(sa$trajectories)
p
ggsave(file.path(OUTPUT_DIR, "sensitivity_trajectories.png"), p,
       width = 10, height = 6, dpi = 120)
```

Three panels (Coal / CoalCap / RenewablesCap) compare production paths
indexed to year-1. The 2025-shock run departs earlier; both converge by 2050.

# 6. Sector x term rollup of PDs

```{r}
pd_by_run <- sa$pd |>
  left_join(sa$params |> select(run_id, shock_year), by = "run_id") |>
  group_by(shock_year, sector, term) |>
  summarise(mean_pd_baseline = mean(pd_baseline, na.rm = TRUE),
            mean_pd_shock    = mean(pd_shock,    na.rm = TRUE),
            .groups = "drop") |>
  arrange(sector, term, shock_year)

knitr::kable(pd_by_run, caption = "Mean PD by sector x term x shock_year") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  kableExtra::scroll_box(width = "100%", height = "400px")

write_csv(pd_by_run, file.path(OUTPUT_DIR, "sensitivity_pd_by_run.csv"))
```
