---
title: "Khan Bank PD/EL Integration Diagnostic"
output:
  html_document:
    toc: true
    toc_depth: 2
    self_contained: true
date: "2026-04-22"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
```

```{r setup}
devtools::load_all(".")
library(ggplot2); library(dplyr); library(tidyr); library(readr)
```

This notebook applies the new **PD/EL integration** methodology (from
`output-development` branch tasks S1+S2) to Khan Bank's 12-firm portfolio.
Every artifact saved to `DEBUG/khanbank-analysis/` is regenerated by knitting
this file.

Khan Bank data lives at `/Users/jakub/Downloads/04.17 debug khanbank/` and is
out of the repo by policy (see `.gitignore` / `.Rbuildignore`).

# 1. Load inputs

```{r}
INPUT_DIR  <- "/Users/jakub/Downloads/04.17 debug khanbank"
OUTPUT_DIR <- "."   # knitting happens in DEBUG/khanbank-analysis/

ad <- readRDS(file.path(INPUT_DIR, "analysis_data.rds")) |>
  mutate(company_id = as.character(company_id))

fin <- read_csv(file.path(INPUT_DIR, "financial_features_final_ifc.csv"),
                show_col_types = FALSE) |>
  select(company_id, bank_internal_pd = pd) |>
  filter(!is.na(company_id)) |>
  mutate(company_id = as.character(company_id))

merton <- read_csv(file.path(INPUT_DIR, "merton_state.csv"),
                   show_col_types = FALSE) |>
  mutate(company_id = as.character(company_id))
```

Drop the row where upstream Merton produced NA PDs (portfolio `term=7`
outside the installed `trisk.model`'s `1:5` grid). The `trisk.model`
commit `5ef6d1b` extends this to `1:10`; reinstalling resolves firm 101.

```{r}
ad_clean <- ad |> filter(!is.na(pd_baseline) & !is.na(pd_shock))
ad_enriched <- compute_analysis_metrics(ad_clean)
cat("Kept", nrow(ad_clean), "of", nrow(ad), "rows\n")
```

# 2. Run the three PD integration methods

```{r}
r_abs <- integrate_pd(ad_enriched, method = "absolute")
r_rel <- integrate_pd(ad_enriched, method = "relative")
r_zs  <- integrate_pd(ad_enriched, method = "zscore")
r_el  <- integrate_el(ad_enriched, method = "relative")
```

# 3. Portfolio KPIs

## 3a. PD KPIs (zscore method)

```{r}
pipeline_crispy_pd_kpi_table(r_zs$aggregate)
```

```{r, include = FALSE}
writeLines(as.character(pipeline_crispy_pd_kpi_table(r_zs$aggregate)),
           file.path(OUTPUT_DIR, "P3a_pd_kpi_table.html"))
```

## 3b. EL KPIs (relative method)

```{r}
pipeline_crispy_el_kpi_table(r_el$aggregate)
```

```{r, include = FALSE}
writeLines(as.character(pipeline_crispy_el_kpi_table(r_el$aggregate)),
           file.path(OUTPUT_DIR, "P3b_el_kpi_table.html"))
```

# 4. Sector breakdown (P4)

```{r}
pipeline_crispy_el_sector_breakdown_table(r_el$portfolio)
```

```{r, include = FALSE}
writeLines(as.character(pipeline_crispy_el_sector_breakdown_table(r_el$portfolio)),
           file.path(OUTPUT_DIR, "P4_el_sector_breakdown.html"))
```

# 5. Visualizations

## P1 — PD integration bars (4-bar grouped per sector x term)

```{r fig.width = 9, fig.height = 6}
p1 <- pipeline_crispy_pd_integration_bars(r_zs)
p1
ggsave(file.path(OUTPUT_DIR, "P1_pd_integration_bars.png"), p1,
       width = 9, height = 6, dpi = 120)
```

## P2 — EL adjustment bars (horizontal, sign-colored)

```{r fig.width = 8, fig.height = 5}
p2 <- pipeline_crispy_el_adjustment_bars(r_el)
p2
ggsave(file.path(OUTPUT_DIR, "P2_el_adjustment_bars.png"), p2,
       width = 8, height = 5, dpi = 120)
```

## N1 — PD method comparison (absolute / relative / zscore)

```{r fig.width = 9, fig.height = 5}
n1 <- pipeline_crispy_pd_method_comparison(ad_enriched)
n1
ggsave(file.path(OUTPUT_DIR, "N1_pd_method_comparison.png"), n1,
       width = 9, height = 5, dpi = 120)
```

# 6. Per-firm diagnostic trace

Each firm gets its pattern classification (`HARD_UNDERFLOW`, `LOW_VOL`,
`LOW_LEVERAGE`, `SHOCK_BENEFIT`, `NEAR_ZERO_PD`, `NORMAL`) plus a
plain-English `why` sentence.

```{r}
m1 <- merton |>
  filter(term == 1) |>
  mutate(
    V0_L_ratio             = (debt + equity_0_baseline) / debt,
    equity_delta_shock_pct = (equity_0_late_sudden - equity_0_baseline) / equity_0_baseline,
    pd_base_raw            = pnorm(-d2_base),
    is_hard_underflow      = d2_base > 37.5
  )

port_term <- ad |>
  select(company_id, technology, port_term = term) |>
  distinct()

diag <- m1 |>
  left_join(port_term, by = "company_id") |>
  left_join(fin, by = "company_id") |>
  mutate(
    pattern = case_when(
      is_hard_underflow             ~ "HARD_UNDERFLOW",
      volatility < 0.06             ~ "LOW_VOL",
      V0_L_ratio > 8                ~ "LOW_LEVERAGE",
      equity_delta_shock_pct > 0.01 ~ "SHOCK_BENEFIT",
      d2_base > 8                   ~ "NEAR_ZERO_PD",
      TRUE                          ~ "NORMAL"
    ),
    narrative = case_when(
      is_hard_underflow             ~ sprintf("d2=%.1f > 37.5 -> pnorm(-d2) = 0 (IEEE 754 underflow)", d2_base),
      volatility < 0.06             ~ sprintf("sigma=%.1f%% low. d2=%.1f -> pnorm(-d2)~10^%d", volatility*100, d2_base, as.integer(log10(pnorm(-d2_base)))),
      V0_L_ratio > 8                ~ sprintf("V0/L=%.1f (low leverage). d2=%.1f -> pnorm(-d2)~10^%d", V0_L_ratio, d2_base, as.integer(log10(pnorm(-d2_base)))),
      equity_delta_shock_pct > 0.01 ~ sprintf("Shock raises equity %+.1f%%. d2_base=%.1f, d2_shock=%.1f, both PDs ~0", equity_delta_shock_pct*100, d2_base, d2_ls),
      d2_base > 8                   ~ sprintf("d2=%.1f -> pnorm(-d2)~10^%d. Below practical precision", d2_base, as.integer(log10(pnorm(-d2_base)))),
      TRUE                          ~ sprintf("d2=%.2f -> PD=%.3f%%. Visible signal", d2_base, pnorm(-d2_base)*100)
    )
  ) |>
  select(company_id, sector, technology, port_term,
         volatility, V0_L_ratio, equity_delta_shock_pct,
         d2_base, d2_ls, pd_base_raw, bank_internal_pd,
         pattern, narrative)

knitr::kable(diag |> select(company_id, sector, pattern, narrative),
             caption = "Per-firm diagnostic narrative") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  kableExtra::scroll_box(width = "100%", height = "600px")

write_csv(diag, file.path(OUTPUT_DIR, "khanbank_pd_diagnostic_perfirm.csv"))
```

# 7. Per-portfolio-row diagnostic with all three methods

```{r}
per_method <- lapply(c("absolute","relative","zscore"), function(m) {
  integrate_pd(ad_enriched, method = m)$portfolio |>
    mutate(company_id = as.character(company_id)) |>
    select(company_id, term, trisk_adjusted_pd) |>
    rename(!!paste0("adjusted_", m) := trisk_adjusted_pd)
})
int_cmp <- Reduce(function(x, y) left_join(x, y, by = c("company_id","term")), per_method)

wide <- ad |>
  select(company_id, sector, technology, term,
         exposure_value_usd, loss_given_default, pd_baseline, pd_shock) |>
  left_join(fin, by = "company_id") |>
  left_join(
    merton |> select(company_id, term, equity_0_baseline, equity_0_late_sudden,
                     debt, d2_base, d2_ls, volatility),
    by = c("company_id", "term")
  ) |>
  mutate(
    V0_L_ratio             = (debt + equity_0_baseline) / debt,
    equity_delta_shock_pct = (equity_0_late_sudden - equity_0_baseline) / equity_0_baseline,
    pd_pattern = case_when(
      is.na(pd_baseline)                ~ "TERM_OUT_OF_MODEL_RANGE",
      !is.na(d2_base) & d2_base > 37.5  ~ "HARD_UNDERFLOW",
      volatility < 0.06                 ~ "LOW_VOL",
      V0_L_ratio > 8                    ~ "LOW_LEVERAGE",
      equity_delta_shock_pct > 0.01     ~ "SHOCK_BENEFIT",
      d2_base > 8                       ~ "NEAR_ZERO_PD",
      TRUE                              ~ "NORMAL"
    )
  ) |>
  left_join(int_cmp, by = c("company_id", "term")) |>
  arrange(company_id, term)

write_csv(wide, file.path(OUTPUT_DIR, "khanbank_pd_diagnostic.csv"))
```
