---
title: "Khan Bank Portfolio Analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
    self_contained: true
date: "2026-04-22"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
```

```{r setup}
devtools::load_all(".")
library(ggplot2); library(dplyr); library(readr)
```

Mirrors `vignettes/portfolio-analysis.Rmd` applied to Khan Bank inputs.
Produces the four standard portfolio plots (NPV change, exposure change, PD
term, expected loss) plus a summary table.

# 1. Load Khan Bank analysis data

The cached `analysis_data.rds` is the output of an earlier
`run_trisk_on_portfolio()` call; we reuse it here to avoid re-running the
full model.

```{r}
INPUT_DIR  <- "/Users/jakub/Downloads/04.17 debug khanbank"
OUTPUT_DIR <- "."

ad <- readRDS(file.path(INPUT_DIR, "analysis_data.rds")) |>
  mutate(company_id = as.character(company_id)) |>
  filter(!is.na(pd_baseline) & !is.na(pd_shock))

cat("rows:", nrow(ad), " sectors:", paste(unique(ad$sector), collapse = ", "), "\n")
```

# 2. Equities risk

## 2a. Average NPV change percentage per technology

```{r fig.width = 9, fig.height = 5}
p1 <- pipeline_crispy_npv_change_plot(ad)
p1
ggsave(file.path(OUTPUT_DIR, "portfolio_npv_change.png"), p1,
       width = 9, height = 5, dpi = 120)
```

Coal and CoalCap techs lose ~100% NPV under the NZ2050 shock; RenewablesCap
gains modestly.

## 2b. Portfolio exposure change

```{r fig.width = 9, fig.height = 5}
p2 <- pipeline_crispy_exposure_change_plot(ad)
p2
ggsave(file.path(OUTPUT_DIR, "portfolio_exposure_change.png"), p2,
       width = 9, height = 5, dpi = 120)
```

# 3. Bonds & Loans risk

## 3a. Average PDs at baseline and shock

```{r fig.width = 9, fig.height = 5}
p3 <- pipeline_crispy_pd_term_plot(ad)
p3
ggsave(file.path(OUTPUT_DIR, "portfolio_pd_term.png"), p3,
       width = 9, height = 5, dpi = 120)
```

Coal shock PDs reach 16% at terms 1 and 4. Power baseline/shock PDs sit at
fractions of a percent, reflecting the low-vol / high-V0/L mechanism
diagnosed in `run_diagnostic.Rmd`.

## 3b. Expected loss by sector

```{r fig.width = 9, fig.height = 5}
p4 <- pipeline_crispy_expected_loss_plot(ad)
p4
ggsave(file.path(OUTPUT_DIR, "portfolio_expected_loss.png"), p4,
       width = 9, height = 5, dpi = 120)
```

Only Coal's shock EL exceeds baseline by a visible amount (thin red loss
bar at the top of the shock column).

# 4. Portfolio summary table

Per firm x term, with NPV, PD, and EL columns used by the plots above.

```{r}
summary_tbl <- compute_analysis_metrics(ad) |>
  select(company_id, company_name, sector, technology, term,
         exposure_value_usd, loss_given_default,
         net_present_value_baseline, net_present_value_shock,
         net_present_value_difference,
         pd_baseline, pd_shock,
         expected_loss_baseline, expected_loss_shock,
         expected_loss_difference) |>
  arrange(company_id, term)

knitr::kable(summary_tbl, caption = "Khan Bank portfolio summary") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  kableExtra::scroll_box(width = "100%", height = "400px")

write_csv(summary_tbl, file.path(OUTPUT_DIR, "portfolio_summary.csv"))
```
